---
  title: "Do GAM to predict Skipjack Tuna biomass from key environmental variables"
output: html_document
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

For quantifying the effect size and direction, we consider a Generalized Additive Model (GAM) approach. This allows us to interpret the coefficients and explanatory power of the parameters, including linear and nonlinear effects.

```{r setup, include=FALSE}

## 1) remove all objects (incl. hidden)
rm(list = ls(all.names = TRUE))
## 2) unload all non-base packages
pkgs <- names(sessionInfo()$otherPkgs)
if (length(pkgs)) sapply(paste0("package:", pkgs),
                         detach, character.only = TRUE,
                         unload = TRUE, force = TRUE)
## 3) clear plots, console, connections, memory
cat("\014")              # clear console in RStudio
gc()


```

```{r}

setwd("/g/data/es60/SEAPODYM/pearse_GAMs")
list.files()

```

Load the dataframes and transform biomass

```{r}

abso <- read.table("/g/data/es60/SEAPODYM/pearse_GAMs/absolute_values_SEAPODYM_WCPO.txt", header = TRUE, sep = ",")
anom <- read.table("/g/data/es60/SEAPODYM/pearse_GAMs/anomalous_values_SEAPODYM_WCPO.txt", header = TRUE, sep = ",")

abso$larvae_biomass_TF <- log10(abs(abso$larvae_biomass)+1.0) * sign(abso$larvae_biomass)
anom$larvae_biomass_TF <- log10(abs(anom$larvae_biomass)+1.0) * sign(anom$larvae_biomass)

abso$juvenile_biomass_TF <- log10(abs(abso$juvenile_biomass)+1.0) * sign(abso$juvenile_biomass)
anom$juvenile_biomass_TF <- log10(abs(anom$juvenile_biomass)+1.0) * sign(anom$juvenile_biomass)

abso$adult_biomass_TF <- log10(abs(abso$adult_biomass)+1.0) * sign(abso$adult_biomass)
anom$adult_biomass_TF <- log10(abs(anom$adult_biomass)+1.0) * sign(anom$adult_biomass)

```

Load libraries

```{r}

library(mgcv)
library(parallel)

```


Create cluster for parallel

```{r}

if (exists("cl")) try(stopCluster(cl), silent = TRUE)

cl <- makeCluster(56)      # or a bit fewer to leave headroom
on.exit(stopCluster(cl), add = TRUE)
length(cl)                 # should be 56


```

Load GAMs

```{r}

gam_abso_larvaeTF_all <- readRDS("gams/gam_abso_larvaeTF_all.rds")
gam_anom_larvaeTF_all <- readRDS("gams/gam_anom_larvaeTF_all.rds")

```

Juveniles Load GAMs

```{r}

gam_abso_juvenileTF_all <- readRDS("gams/gam_abso_juvenileTF_all.rds")
gam_abso_juvenileTF_allL <- readRDS("gams/gam_abso_juvenileTF_allL.rds")
gam_anom_juvenileTF_all <- readRDS("gams/gam_anom_juvenileTF_all.rds")
gam_anom_juvenileTF_allL <- readRDS("gams/gam_anom_juvenileTF_allL.rds")
gam_abso_juvenileTF_allpL <- readRDS("gams/gam_abso_juvenileTF_allpL.rds")
gam_anom_juvenileTF_allpL <- readRDS("gams/gam_anom_juvenileTF_allpL.rds")

```

Predicted larvae

```{r}

abso$larvae_biomass_TFpred_all <- predict(gam_abso_larvaeTF_all, newdata=abso)
anom$larvae_biomass_TFpred_all <- predict(gam_anom_larvaeTF_all, newdata=anom)

```

Adults

```{r}

gam_abso_adultTF_all <- readRDS("gams/gam_abso_adultTF_all.rds")
gam_abso_adultTF_allJ <- readRDS("gams/gam_abso_adultTF_allJ.rds")
gam_anom_adultTF_all <- readRDS("gams/gam_anom_adultTF_all.rds")
gam_anom_adultTF_allJ <- readRDS("gams/gam_anom_adultTF_allJ.rds")
gam_abso_adultTF_allpL <- readRDS("gams/gam_abso_adultTF_allpJ.rds")
gam_anom_adultTF_allpL <- readRDS("gams/gam_anom_adultTF_allpJ.rds")

```

Predicted juveniles 

```{r}

abso$juvenile_biomass_TFpred_all <- predict(gam_abso_juvenileTF_allpL, newdata=abso)
anom$juvenile_biomass_TFpred_all <- predict(gam_anom_juvenileTF_allpL, newdata=anom)

```

####################################################
Assess model skills using a test-train split 80%-20%
####################################################

First, sort the data according to season_year

```{r}

abso_sorted <- abso[ order(abso$season_year), ]
anom_sorted <- anom[ order(anom$season_year), ]

```

```{r}

# total rows
N <- nrow(anom_sorted)

# 80% cutoff row
cut_row <- floor(0.8 * N)

# 3. Create the splits
abso_train <- abso_sorted[1:cut_row, ]
abso_test  <- abso_sorted[(cut_row + 1):N, ]
anom_train <- anom_sorted[1:cut_row, ]
anom_test  <- anom_sorted[(cut_row + 1):N, ]

anom_test

```


```{r}

k1 <- 5

gam80_abso_larvaeTF_all <- bam(
  larvae_biomass_TF ~ 
    # main effects (nonlinear, global)
    s(sst, k=k1, bs="ts") + 
    s(mld, k=k1, bs="ts") +
    s(zeu, k=k1, bs="ts") +
    s(uo, k=k1, bs="ts") +
    s(vo, k=k1, bs="ts") +
    s(pp, k=k1, bs="ts") +
    s(T_L1, k=k1, bs="ts") +
    s(T_L2, k=k1, bs="ts") +
    s(T_L3, k=k1, bs="ts") +
    s(U_L1, k=k1, bs="ts") +
    s(U_L2, k=k1, bs="ts") +
    s(U_L3, k=k1, bs="ts") +
    s(V_L1, k=k1, bs="ts") +
    s(V_L2, k=k1, bs="ts") +
    s(V_L3, k=k1, bs="ts") +
    s(O2_L1, k=k1, bs="ts") +
    s(O2_L2, k=k1, bs="ts") +
    s(O2_L3, k=k1, bs="ts") +
    # space-varying effects
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(mld, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(zeu, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(uo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(vo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = abso_train,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam80_abso_larvaeTF_all)

gam80_anom_larvaeTF_all <- bam(
  larvae_biomass_TF ~ 
    # main effects (nonlinear, global)
    s(sst, k=k1, bs="ts") + 
    s(mld, k=k1, bs="ts") +
    s(zeu, k=k1, bs="ts") +
    s(uo, k=k1, bs="ts") +
    s(vo, k=k1, bs="ts") +
    s(pp, k=k1, bs="ts") +
    s(T_L1, k=k1, bs="ts") +
    s(T_L2, k=k1, bs="ts") +
    s(T_L3, k=k1, bs="ts") +
    s(U_L1, k=k1, bs="ts") +
    s(U_L2, k=k1, bs="ts") +
    s(U_L3, k=k1, bs="ts") +
    s(V_L1, k=k1, bs="ts") +
    s(V_L2, k=k1, bs="ts") +
    s(V_L3, k=k1, bs="ts") +
    s(O2_L1, k=k1, bs="ts") +
    s(O2_L2, k=k1, bs="ts") +
    s(O2_L3, k=k1, bs="ts") +
    # space-varying effects
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(mld, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(zeu, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(uo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(vo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom_train,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam80_anom_larvaeTF_all)

abso_clim_train <- tapply(  abso_train$larvae_biomass_TF,  abso_train$season,  mean,  na.rm = TRUE )
abso_clim_test <- abso_clim_train[ abso_test$season ]
anom_clim_train <- tapply(  anom_train$larvae_biomass_TF,  anom_train$season,  mean,  na.rm = TRUE )
anom_clim_test <- anom_clim_train[ anom_test$season ]

pred_abso_all <- predict(gam80_abso_larvaeTF_all, newdata = abso_test)
pred_anom_all <- predict(gam80_anom_larvaeTF_all, newdata = anom_test)

idx <- complete.cases(abso_test$larvae_biomass_TF, pred_abso_all, abso_test$season, abso_clim_test)
y_test   <- abso_test$larvae_biomass_TF[idx]
y_pred   <- pred_abso_all[idx]
clim_use <- abso_clim_test[idx]
SS_res <- sum((y_test - y_pred)^2)
SS_tot <- sum((y_test - clim_use)^2)
R2_abso_larvae_all <- 1 - SS_res / SS_tot

idx <- complete.cases(anom_test$larvae_biomass_TF, pred_anom_all, anom_test$season, anom_clim_test)
y_test   <- anom_test$larvae_biomass_TF[idx]
y_pred   <- pred_anom_all[idx]
clim_use <- anom_clim_test[idx]
SS_res <- sum((y_test - y_pred)^2)
SS_tot <- sum((y_test - clim_use)^2)
R2_anom_larvae_all <- 1 - SS_res / SS_tot

print(R2_abso_larvae_all)
print(R2_anom_larvae_all)

```

Juveniles (no larvae)

```{r}

k1 <- 5

gam80_abso_juvenileTF_all <- bam(
  juvenile_biomass_TF ~ 
    # main effects (nonlinear, global)
    s(sst, k=k1, bs="ts") + 
    s(mld, k=k1, bs="ts") +
    s(zeu, k=k1, bs="ts") +
    s(uo, k=k1, bs="ts") +
    s(vo, k=k1, bs="ts") +
    s(pp, k=k1, bs="ts") +
    s(T_L1, k=k1, bs="ts") +
    s(T_L2, k=k1, bs="ts") +
    s(T_L3, k=k1, bs="ts") +
    s(U_L1, k=k1, bs="ts") +
    s(U_L2, k=k1, bs="ts") +
    s(U_L3, k=k1, bs="ts") +
    s(V_L1, k=k1, bs="ts") +
    s(V_L2, k=k1, bs="ts") +
    s(V_L3, k=k1, bs="ts") +
    s(O2_L1, k=k1, bs="ts") +
    s(O2_L2, k=k1, bs="ts") +
    s(O2_L3, k=k1, bs="ts") +
    # space-varying effects
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(mld, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(zeu, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(uo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(vo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = abso_train,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam80_abso_juvenileTF_all)

gam80_anom_juvenileTF_all <- bam(
  juvenile_biomass_TF ~ 
    # main effects (nonlinear, global)
    s(sst, k=k1, bs="ts") + 
    s(mld, k=k1, bs="ts") +
    s(zeu, k=k1, bs="ts") +
    s(uo, k=k1, bs="ts") +
    s(vo, k=k1, bs="ts") +
    s(pp, k=k1, bs="ts") +
    s(T_L1, k=k1, bs="ts") +
    s(T_L2, k=k1, bs="ts") +
    s(T_L3, k=k1, bs="ts") +
    s(U_L1, k=k1, bs="ts") +
    s(U_L2, k=k1, bs="ts") +
    s(U_L3, k=k1, bs="ts") +
    s(V_L1, k=k1, bs="ts") +
    s(V_L2, k=k1, bs="ts") +
    s(V_L3, k=k1, bs="ts") +
    s(O2_L1, k=k1, bs="ts") +
    s(O2_L2, k=k1, bs="ts") +
    s(O2_L3, k=k1, bs="ts") +
    # space-varying effects
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(mld, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(zeu, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(uo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(vo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom_train,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam80_anom_juvenileTF_all)


abso_clim_train <- tapply(  abso_train$juvenile_biomass_TF,  abso_train$season,  mean,  na.rm = TRUE )
abso_clim_test <- abso_clim_train[ abso_test$season ]
anom_clim_train <- tapply(  anom_train$juvenile_biomass_TF,  anom_train$season,  mean,  na.rm = TRUE )
anom_clim_test <- anom_clim_train[ anom_test$season ]

pred_abso_all <- predict(gam80_abso_juvenileTF_all, newdata = abso_test)
pred_anom_all <- predict(gam80_anom_juvenileTF_all, newdata = anom_test)

idx <- complete.cases(abso_test$juvenile_biomass_TF, pred_abso_all, abso_test$season, abso_clim_test)
y_test   <- abso_test$juvenile_biomass_TF[idx]
y_pred   <- pred_abso_all[idx]
clim_use <- abso_clim_test[idx]
SS_res <- sum((y_test - y_pred)^2)
SS_tot <- sum((y_test - clim_use)^2)
R2_abso_juvenile_all <- 1 - SS_res / SS_tot

idx <- complete.cases(anom_test$juvenile_biomass_TF, pred_anom_all, anom_test$season, anom_clim_test)
y_test   <- anom_test$juvenile_biomass_TF[idx]
y_pred   <- pred_anom_all[idx]
clim_use <- anom_clim_test[idx]
SS_res <- sum((y_test - y_pred)^2)
SS_tot <- sum((y_test - clim_use)^2)
R2_anom_juvenile_all <- 1 - SS_res / SS_tot

print(R2_abso_juvenile_all)
print(R2_anom_juvenile_all)

```

Juveniles with perfectly known larvae

```{r}

k1 <- 5

gam80_abso_juvenileTF_allL <- bam(
  juvenile_biomass_TF ~ 
    # main effects (nonlinear, global)
    s(sst, k=k1, bs="ts") + 
    s(mld, k=k1, bs="ts") +
    s(zeu, k=k1, bs="ts") +
    s(uo, k=k1, bs="ts") +
    s(vo, k=k1, bs="ts") +
    s(pp, k=k1, bs="ts") +
    s(T_L1, k=k1, bs="ts") +
    s(T_L2, k=k1, bs="ts") +
    s(T_L3, k=k1, bs="ts") +
    s(U_L1, k=k1, bs="ts") +
    s(U_L2, k=k1, bs="ts") +
    s(U_L3, k=k1, bs="ts") +
    s(V_L1, k=k1, bs="ts") +
    s(V_L2, k=k1, bs="ts") +
    s(V_L3, k=k1, bs="ts") +
    s(O2_L1, k=k1, bs="ts") +
    s(O2_L2, k=k1, bs="ts") +
    s(O2_L3, k=k1, bs="ts") +
    s(larvae_biomass_TF, k=k1, bs="ts") +
    # space-varying effects
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(mld, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(zeu, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(uo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(vo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(larvae_biomass_TF, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = abso_train,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam80_abso_juvenileTF_allL)

gam80_anom_juvenileTF_allL <- bam(
  juvenile_biomass_TF ~ 
    # main effects (nonlinear, global)
    s(sst, k=k1, bs="ts") + 
    s(mld, k=k1, bs="ts") +
    s(zeu, k=k1, bs="ts") +
    s(uo, k=k1, bs="ts") +
    s(vo, k=k1, bs="ts") +
    s(pp, k=k1, bs="ts") +
    s(T_L1, k=k1, bs="ts") +
    s(T_L2, k=k1, bs="ts") +
    s(T_L3, k=k1, bs="ts") +
    s(U_L1, k=k1, bs="ts") +
    s(U_L2, k=k1, bs="ts") +
    s(U_L3, k=k1, bs="ts") +
    s(V_L1, k=k1, bs="ts") +
    s(V_L2, k=k1, bs="ts") +
    s(V_L3, k=k1, bs="ts") +
    s(O2_L1, k=k1, bs="ts") +
    s(O2_L2, k=k1, bs="ts") +
    s(O2_L3, k=k1, bs="ts") +
    s(larvae_biomass_TF, k=k1, bs="ts") +
    # space-varying effects
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(mld, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(zeu, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(uo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(vo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(larvae_biomass_TF, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom_train,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam80_anom_juvenileTF_allL)

abso_clim_train <- tapply(  abso_train$juvenile_biomass_TF,  abso_train$season,  mean,  na.rm = TRUE )
abso_clim_test <- abso_clim_train[ abso_test$season ]
anom_clim_train <- tapply(  anom_train$juvenile_biomass_TF,  anom_train$season,  mean,  na.rm = TRUE )
anom_clim_test <- anom_clim_train[ anom_test$season ]

pred_abso_allL <- predict(gam80_abso_juvenileTF_allL, newdata = abso_test)
pred_anom_allL <- predict(gam80_anom_juvenileTF_allL, newdata = anom_test)

idx <- complete.cases(abso_test$juvenile_biomass_TF, pred_abso_allL, abso_test$season, abso_clim_test)
y_test   <- abso_test$juvenile_biomass_TF[idx]
y_pred   <- pred_abso_allL[idx]
clim_use <- abso_clim_test[idx]
SS_res <- sum((y_test - y_pred)^2)
SS_tot <- sum((y_test - clim_use)^2)
R2_abso_juvenile_allL <- 1 - SS_res / SS_tot

idx <- complete.cases(anom_test$juvenile_biomass_TF, pred_anom_allL, anom_test$season, anom_clim_test)
y_test   <- anom_test$juvenile_biomass_TF[idx]
y_pred   <- pred_anom_allL[idx]
clim_use <- anom_clim_test[idx]
SS_res <- sum((y_test - y_pred)^2)
SS_tot <- sum((y_test - clim_use)^2)
R2_anom_juvenile_allL <- 1 - SS_res / SS_tot

print(R2_abso_juvenile_allL)
print(R2_anom_juvenile_allL)

```

Juveniles with predicted larvae

```{r}

k1 <- 5

gam80_abso_juvenileTF_allpL <- bam(
  juvenile_biomass_TF ~ 
    # main effects (nonlinear, global)
    s(sst, k=k1, bs="ts") + 
    s(mld, k=k1, bs="ts") +
    s(zeu, k=k1, bs="ts") +
    s(uo, k=k1, bs="ts") +
    s(vo, k=k1, bs="ts") +
    s(pp, k=k1, bs="ts") +
    s(T_L1, k=k1, bs="ts") +
    s(T_L2, k=k1, bs="ts") +
    s(T_L3, k=k1, bs="ts") +
    s(U_L1, k=k1, bs="ts") +
    s(U_L2, k=k1, bs="ts") +
    s(U_L3, k=k1, bs="ts") +
    s(V_L1, k=k1, bs="ts") +
    s(V_L2, k=k1, bs="ts") +
    s(V_L3, k=k1, bs="ts") +
    s(O2_L1, k=k1, bs="ts") +
    s(O2_L2, k=k1, bs="ts") +
    s(O2_L3, k=k1, bs="ts") +
    s(larvae_biomass_TFpred_all, k=k1, bs="ts") +
    # space-varying effects
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(mld, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(zeu, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(uo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(vo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(larvae_biomass_TFpred_all, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = abso_train,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam80_abso_juvenileTF_allpL)

gam80_anom_juvenileTF_allpL <- bam(
  juvenile_biomass_TF ~ 
    # main effects (nonlinear, global)
    s(sst, k=k1, bs="ts") + 
    s(mld, k=k1, bs="ts") +
    s(zeu, k=k1, bs="ts") +
    s(uo, k=k1, bs="ts") +
    s(vo, k=k1, bs="ts") +
    s(pp, k=k1, bs="ts") +
    s(T_L1, k=k1, bs="ts") +
    s(T_L2, k=k1, bs="ts") +
    s(T_L3, k=k1, bs="ts") +
    s(U_L1, k=k1, bs="ts") +
    s(U_L2, k=k1, bs="ts") +
    s(U_L3, k=k1, bs="ts") +
    s(V_L1, k=k1, bs="ts") +
    s(V_L2, k=k1, bs="ts") +
    s(V_L3, k=k1, bs="ts") +
    s(O2_L1, k=k1, bs="ts") +
    s(O2_L2, k=k1, bs="ts") +
    s(O2_L3, k=k1, bs="ts") +
    s(larvae_biomass_TFpred_all, k=k1, bs="ts") +
    # space-varying effects
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(mld, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(zeu, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(uo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(vo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(larvae_biomass_TFpred_all, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom_train,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam80_anom_juvenileTF_allpL)

abso_clim_train <- tapply(  abso_train$juvenile_biomass_TF,  abso_train$season,  mean,  na.rm = TRUE )
abso_clim_test <- abso_clim_train[ abso_test$season ]
anom_clim_train <- tapply(  anom_train$juvenile_biomass_TF,  anom_train$season,  mean,  na.rm = TRUE )
anom_clim_test <- anom_clim_train[ anom_test$season ]

pred_abso_allpL <- predict(gam80_abso_juvenileTF_allpL, newdata = abso_test)
pred_anom_allpL <- predict(gam80_anom_juvenileTF_allpL, newdata = anom_test)

idx <- complete.cases(abso_test$juvenile_biomass_TF, pred_abso_allpL, abso_test$season, abso_clim_test)
y_test   <- abso_test$juvenile_biomass_TF[idx]
y_pred   <- pred_abso_allpL[idx]
clim_use <- abso_clim_test[idx]
SS_res <- sum((y_test - y_pred)^2)
SS_tot <- sum((y_test - clim_use)^2)
R2_abso_juvenile_allpL <- 1 - SS_res / SS_tot

idx <- complete.cases(anom_test$juvenile_biomass_TF, pred_anom_allpL, anom_test$season, anom_clim_test)
y_test   <- anom_test$juvenile_biomass_TF[idx]
y_pred   <- pred_anom_allpL[idx]
clim_use <- anom_clim_test[idx]
SS_res <- sum((y_test - y_pred)^2)
SS_tot <- sum((y_test - clim_use)^2)
R2_anom_juvenile_allpL <- 1 - SS_res / SS_tot

print(R2_abso_juvenile_allpL)
print(R2_anom_juvenile_allpL)

```

Juveniles just from larvae

```{r}

gam80_abso_juvenileTF_J <- bam(
  juvenile_biomass_TF ~ 
    s(larvae_biomass_TF, k=k1, bs="ts") +
    ti(larvae_biomass_TF, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = abso_train,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam80_abso_juvenileTF_J)

gam80_abso_juvenileTF_pJ <- bam(
  juvenile_biomass_TF ~ 
    s(larvae_biomass_TFpred_all, k=k1, bs="ts") +
    ti(larvae_biomass_TFpred_all, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = abso_train,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam80_abso_juvenileTF_pJ)

gam80_anom_juvenileTF_J <- bam(
  juvenile_biomass_TF ~ 
    s(larvae_biomass_TF, k=k1, bs="ts") +
    ti(larvae_biomass_TF, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom_train,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam80_anom_juvenileTF_J)

gam80_anom_juvenileTF_pJ <- bam(
  juvenile_biomass_TF ~ 
    s(larvae_biomass_TFpred_all, k=k1, bs="ts") +
    ti(larvae_biomass_TFpred_all, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom_train,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam80_anom_juvenileTF_pJ)

abso_clim_train <- tapply(  abso_train$juvenile_biomass_TF,  abso_train$season,  mean,  na.rm = TRUE )
abso_clim_test <- abso_clim_train[ abso_test$season ]
anom_clim_train <- tapply(  anom_train$juvenile_biomass_TF,  anom_train$season,  mean,  na.rm = TRUE )
anom_clim_test <- anom_clim_train[ anom_test$season ]

pred_abso_J <- predict(gam80_abso_juvenileTF_J, newdata = abso_test)
pred_abso_pJ <- predict(gam80_abso_juvenileTF_pJ, newdata = abso_test)
pred_anom_J <- predict(gam80_anom_juvenileTF_J, newdata = anom_test)
pred_anom_pJ <- predict(gam80_anom_juvenileTF_pJ, newdata = anom_test)

idx <- complete.cases(abso_test$juvenile_biomass_TF, pred_abso_J, abso_test$season, abso_clim_test)
y_test   <- abso_test$juvenile_biomass_TF[idx]
y_pred   <- pred_abso_J[idx]
clim_use <- abso_clim_test[idx]
SS_res <- sum((y_test - y_pred)^2)
SS_tot <- sum((y_test - clim_use)^2)
R2_abso_juvenile_J <- 1 - SS_res / SS_tot

idx <- complete.cases(abso_test$juvenile_biomass_TF, pred_abso_pJ, abso_test$season, abso_clim_test)
y_test   <- abso_test$juvenile_biomass_TF[idx]
y_pred   <- pred_abso_pJ[idx]
clim_use <- abso_clim_test[idx]
SS_res <- sum((y_test - y_pred)^2)
SS_tot <- sum((y_test - clim_use)^2)
R2_abso_juvenile_pJ <- 1 - SS_res / SS_tot

idx <- complete.cases(anom_test$juvenile_biomass_TF, pred_anom_J, anom_test$season, anom_clim_test)
y_test   <- anom_test$juvenile_biomass_TF[idx]
y_pred   <- pred_anom_J[idx]
clim_use <- anom_clim_test[idx]
SS_res <- sum((y_test - y_pred)^2)
SS_tot <- sum((y_test - clim_use)^2)
R2_anom_juvenile_J <- 1 - SS_res / SS_tot

idx <- complete.cases(anom_test$juvenile_biomass_TF, pred_anom_pJ, anom_test$season, anom_clim_test)
y_test   <- anom_test$juvenile_biomass_TF[idx]
y_pred   <- pred_anom_pJ[idx]
clim_use <- anom_clim_test[idx]
SS_res <- sum((y_test - y_pred)^2)
SS_tot <- sum((y_test - clim_use)^2)
R2_anom_juvenile_pJ <- 1 - SS_res / SS_tot


print(R2_abso_juvenile_J)
print(R2_abso_juvenile_pJ)
print(R2_anom_juvenile_J)
print(R2_anom_juvenile_pJ)

```


Adults (no juveniles)

```{r}

k1 <- 5

gam80_abso_adultTF_all <- bam(
  adult_biomass_TF ~ 
    # main effects (nonlinear, global)
    s(sst, k=k1, bs="ts") + 
    s(mld, k=k1, bs="ts") +
    s(zeu, k=k1, bs="ts") +
    s(uo, k=k1, bs="ts") +
    s(vo, k=k1, bs="ts") +
    s(pp, k=k1, bs="ts") +
    s(T_L1, k=k1, bs="ts") +
    s(T_L2, k=k1, bs="ts") +
    s(T_L3, k=k1, bs="ts") +
    s(U_L1, k=k1, bs="ts") +
    s(U_L2, k=k1, bs="ts") +
    s(U_L3, k=k1, bs="ts") +
    s(V_L1, k=k1, bs="ts") +
    s(V_L2, k=k1, bs="ts") +
    s(V_L3, k=k1, bs="ts") +
    s(O2_L1, k=k1, bs="ts") +
    s(O2_L2, k=k1, bs="ts") +
    s(O2_L3, k=k1, bs="ts") +
    # space-varying effects
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(mld, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(zeu, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(uo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(vo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = abso_train,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam80_abso_adultTF_all)

gam80_anom_adultTF_all <- bam(
  adult_biomass_TF ~ 
    # main effects (nonlinear, global)
    s(sst, k=k1, bs="ts") + 
    s(mld, k=k1, bs="ts") +
    s(zeu, k=k1, bs="ts") +
    s(uo, k=k1, bs="ts") +
    s(vo, k=k1, bs="ts") +
    s(pp, k=k1, bs="ts") +
    s(T_L1, k=k1, bs="ts") +
    s(T_L2, k=k1, bs="ts") +
    s(T_L3, k=k1, bs="ts") +
    s(U_L1, k=k1, bs="ts") +
    s(U_L2, k=k1, bs="ts") +
    s(U_L3, k=k1, bs="ts") +
    s(V_L1, k=k1, bs="ts") +
    s(V_L2, k=k1, bs="ts") +
    s(V_L3, k=k1, bs="ts") +
    s(O2_L1, k=k1, bs="ts") +
    s(O2_L2, k=k1, bs="ts") +
    s(O2_L3, k=k1, bs="ts") +
    # space-varying effects
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(mld, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(zeu, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(uo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(vo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom_train,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam80_anom_adultTF_all)

abso_clim_train <- tapply(  abso_train$adult_biomass_TF,  abso_train$season,  mean,  na.rm = TRUE )
abso_clim_test <- abso_clim_train[ abso_test$season ]
anom_clim_train <- tapply(  anom_train$adult_biomass_TF,  anom_train$season,  mean,  na.rm = TRUE )
anom_clim_test <- anom_clim_train[ anom_test$season ]

pred_abso_all <- predict(gam80_abso_adultTF_all, newdata = abso_test)
pred_anom_all <- predict(gam80_anom_adultTF_all, newdata = anom_test)

idx <- complete.cases(abso_test$adult_biomass_TF, pred_abso_all, abso_test$season, abso_clim_test)
y_test   <- abso_test$adult_biomass_TF[idx]
y_pred   <- pred_abso_all[idx]
clim_use <- abso_clim_test[idx]
SS_res <- sum((y_test - y_pred)^2)
SS_tot <- sum((y_test - clim_use)^2)
R2_abso_adult_all <- 1 - SS_res / SS_tot

idx <- complete.cases(anom_test$adult_biomass_TF, pred_anom_all, anom_test$season, anom_clim_test)
y_test   <- anom_test$adult_biomass_TF[idx]
y_pred   <- pred_anom_all[idx]
clim_use <- anom_clim_test[idx]
SS_res <- sum((y_test - y_pred)^2)
SS_tot <- sum((y_test - clim_use)^2)
R2_anom_adult_all <- 1 - SS_res / SS_tot

print(R2_abso_adult_all)
print(R2_anom_adult_all)

```

Adults with perfectly known juveniles

```{r}

k1 <- 5

gam80_abso_adultTF_allJ <- bam(
  adult_biomass_TF ~ 
    # main effects (nonlinear, global)
    s(sst, k=k1, bs="ts") + 
    s(mld, k=k1, bs="ts") +
    s(zeu, k=k1, bs="ts") +
    s(uo, k=k1, bs="ts") +
    s(vo, k=k1, bs="ts") +
    s(pp, k=k1, bs="ts") +
    s(T_L1, k=k1, bs="ts") +
    s(T_L2, k=k1, bs="ts") +
    s(T_L3, k=k1, bs="ts") +
    s(U_L1, k=k1, bs="ts") +
    s(U_L2, k=k1, bs="ts") +
    s(U_L3, k=k1, bs="ts") +
    s(V_L1, k=k1, bs="ts") +
    s(V_L2, k=k1, bs="ts") +
    s(V_L3, k=k1, bs="ts") +
    s(O2_L1, k=k1, bs="ts") +
    s(O2_L2, k=k1, bs="ts") +
    s(O2_L3, k=k1, bs="ts") +
    s(juvenile_biomass_TF, k=k1, bs="ts") +
    # space-varying effects
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(mld, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(zeu, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(uo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(vo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(juvenile_biomass_TF, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = abso_train,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam80_abso_adultTF_allJ)

gam80_anom_adultTF_allJ <- bam(
  adult_biomass_TF ~ 
    # main effects (nonlinear, global)
    s(sst, k=k1, bs="ts") + 
    s(mld, k=k1, bs="ts") +
    s(zeu, k=k1, bs="ts") +
    s(uo, k=k1, bs="ts") +
    s(vo, k=k1, bs="ts") +
    s(pp, k=k1, bs="ts") +
    s(T_L1, k=k1, bs="ts") +
    s(T_L2, k=k1, bs="ts") +
    s(T_L3, k=k1, bs="ts") +
    s(U_L1, k=k1, bs="ts") +
    s(U_L2, k=k1, bs="ts") +
    s(U_L3, k=k1, bs="ts") +
    s(V_L1, k=k1, bs="ts") +
    s(V_L2, k=k1, bs="ts") +
    s(V_L3, k=k1, bs="ts") +
    s(O2_L1, k=k1, bs="ts") +
    s(O2_L2, k=k1, bs="ts") +
    s(O2_L3, k=k1, bs="ts") +
    s(juvenile_biomass_TF, k=k1, bs="ts") +
    # space-varying effects
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(mld, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(zeu, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(uo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(vo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(juvenile_biomass_TF, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom_train,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam80_anom_adultTF_allJ)

abso_clim_train <- tapply(  abso_train$adult_biomass_TF,  abso_train$season,  mean,  na.rm = TRUE )
abso_clim_test <- abso_clim_train[ abso_test$season ]
anom_clim_train <- tapply(  anom_train$adult_biomass_TF,  anom_train$season,  mean,  na.rm = TRUE )
anom_clim_test <- anom_clim_train[ anom_test$season ]

pred_abso_allJ <- predict(gam80_abso_adultTF_allJ, newdata = abso_test)
pred_anom_allJ <- predict(gam80_anom_adultTF_allJ, newdata = anom_test)

idx <- complete.cases(abso_test$adult_biomass_TF, pred_abso_allJ, abso_test$season, abso_clim_test)
y_test   <- abso_test$adult_biomass_TF[idx]
y_pred   <- pred_abso_allJ[idx]
clim_use <- abso_clim_test[idx]
SS_res <- sum((y_test - y_pred)^2)
SS_tot <- sum((y_test - clim_use)^2)
R2_abso_adult_allJ <- 1 - SS_res / SS_tot

idx <- complete.cases(anom_test$adult_biomass_TF, pred_anom_allJ, anom_test$season, anom_clim_test)
y_test   <- anom_test$adult_biomass_TF[idx]
y_pred   <- pred_anom_allJ[idx]
clim_use <- anom_clim_test[idx]
SS_res <- sum((y_test - y_pred)^2)
SS_tot <- sum((y_test - clim_use)^2)
R2_anom_adult_allJ <- 1 - SS_res / SS_tot

print(R2_abso_adult_allJ)
print(R2_anom_adult_allJ)

```

Adults with perfectly known juveniles

```{r}

k1 <- 5

gam80_abso_adultTF_allpJ <- bam(
  adult_biomass_TF ~ 
    # main effects (nonlinear, global)
    s(sst, k=k1, bs="ts") + 
    s(mld, k=k1, bs="ts") +
    s(zeu, k=k1, bs="ts") +
    s(uo, k=k1, bs="ts") +
    s(vo, k=k1, bs="ts") +
    s(pp, k=k1, bs="ts") +
    s(T_L1, k=k1, bs="ts") +
    s(T_L2, k=k1, bs="ts") +
    s(T_L3, k=k1, bs="ts") +
    s(U_L1, k=k1, bs="ts") +
    s(U_L2, k=k1, bs="ts") +
    s(U_L3, k=k1, bs="ts") +
    s(V_L1, k=k1, bs="ts") +
    s(V_L2, k=k1, bs="ts") +
    s(V_L3, k=k1, bs="ts") +
    s(O2_L1, k=k1, bs="ts") +
    s(O2_L2, k=k1, bs="ts") +
    s(O2_L3, k=k1, bs="ts") +
    s(juvenile_biomass_TFpred_all, k=k1, bs="ts") +
    # space-varying effects
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(mld, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(zeu, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(uo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(vo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(juvenile_biomass_TFpred_all, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = abso_train,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam80_abso_adultTF_allpJ)

gam80_anom_adultTF_allpJ <- bam(
  adult_biomass_TF ~ 
    # main effects (nonlinear, global)
    s(sst, k=k1, bs="ts") + 
    s(mld, k=k1, bs="ts") +
    s(zeu, k=k1, bs="ts") +
    s(uo, k=k1, bs="ts") +
    s(vo, k=k1, bs="ts") +
    s(pp, k=k1, bs="ts") +
    s(T_L1, k=k1, bs="ts") +
    s(T_L2, k=k1, bs="ts") +
    s(T_L3, k=k1, bs="ts") +
    s(U_L1, k=k1, bs="ts") +
    s(U_L2, k=k1, bs="ts") +
    s(U_L3, k=k1, bs="ts") +
    s(V_L1, k=k1, bs="ts") +
    s(V_L2, k=k1, bs="ts") +
    s(V_L3, k=k1, bs="ts") +
    s(O2_L1, k=k1, bs="ts") +
    s(O2_L2, k=k1, bs="ts") +
    s(O2_L3, k=k1, bs="ts") +
    s(juvenile_biomass_TFpred_all, k=k1, bs="ts") +
    # space-varying effects
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(mld, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(zeu, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(uo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(vo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(juvenile_biomass_TFpred_all, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom_train,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam80_anom_adultTF_allpJ)

abso_clim_train <- tapply(  abso_train$adult_biomass_TF,  abso_train$season,  mean,  na.rm = TRUE )
abso_clim_test <- abso_clim_train[ abso_test$season ]
anom_clim_train <- tapply(  anom_train$adult_biomass_TF,  anom_train$season,  mean,  na.rm = TRUE )
anom_clim_test <- anom_clim_train[ anom_test$season ]

pred_abso_allpJ <- predict(gam80_abso_adultTF_allpJ, newdata = abso_test)
pred_anom_allpJ <- predict(gam80_anom_adultTF_allpJ, newdata = anom_test)

idx <- complete.cases(abso_test$adult_biomass_TF, pred_abso_allpJ, abso_test$season, abso_clim_test)
y_test   <- abso_test$adult_biomass_TF[idx]
y_pred   <- pred_abso_allpJ[idx]
clim_use <- abso_clim_test[idx]
SS_res <- sum((y_test - y_pred)^2)
SS_tot <- sum((y_test - clim_use)^2)
R2_abso_adult_allpJ <- 1 - SS_res / SS_tot

idx <- complete.cases(anom_test$adult_biomass_TF, pred_anom_allpJ, anom_test$season, anom_clim_test)
y_test   <- anom_test$adult_biomass_TF[idx]
y_pred   <- pred_anom_allpJ[idx]
clim_use <- anom_clim_test[idx]
SS_res <- sum((y_test - y_pred)^2)
SS_tot <- sum((y_test - clim_use)^2)
R2_anom_adult_allpJ <- 1 - SS_res / SS_tot


print(R2_abso_adult_allpJ)
print(R2_anom_adult_allpJ)

```

```{r}

gam80_abso_adultTF_J <- bam(
  adult_biomass_TF ~ 
    s(juvenile_biomass_TF, k=k1, bs="ts") +
    ti(juvenile_biomass_TF, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = abso_train,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam80_abso_adultTF_J)

gam80_abso_adultTF_pJ <- bam(
  adult_biomass_TF ~ 
    s(juvenile_biomass_TFpred_all, k=k1, bs="ts") +
    ti(juvenile_biomass_TFpred_all, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = abso_train,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam80_abso_adultTF_pJ)

gam80_anom_adultTF_J <- bam(
  adult_biomass_TF ~ 
    s(juvenile_biomass_TF, k=k1, bs="ts") +
    ti(juvenile_biomass_TF, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom_train,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam80_anom_adultTF_J)

gam80_anom_adultTF_pJ <- bam(
  adult_biomass_TF ~ 
    s(juvenile_biomass_TFpred_all, k=k1, bs="ts") +
    ti(juvenile_biomass_TFpred_all, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom_train,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam80_anom_adultTF_pJ)

abso_clim_train <- tapply(  abso_train$adult_biomass_TF,  abso_train$season,  mean,  na.rm = TRUE )
abso_clim_test <- abso_clim_train[ abso_test$season ]
anom_clim_train <- tapply(  anom_train$adult_biomass_TF,  anom_train$season,  mean,  na.rm = TRUE )
anom_clim_test <- anom_clim_train[ anom_test$season ]

pred_abso_J <- predict(gam80_abso_adultTF_J, newdata = abso_test)
pred_abso_pJ <- predict(gam80_abso_adultTF_pJ, newdata = abso_test)
pred_anom_J <- predict(gam80_anom_adultTF_J, newdata = anom_test)
pred_anom_pJ <- predict(gam80_anom_adultTF_pJ, newdata = anom_test)

idx <- complete.cases(abso_test$adult_biomass_TF, pred_abso_J, abso_test$season, abso_clim_test)
y_test   <- abso_test$adult_biomass_TF[idx]
y_pred   <- pred_abso_J[idx]
clim_use <- abso_clim_test[idx]
SS_res <- sum((y_test - y_pred)^2)
SS_tot <- sum((y_test - clim_use)^2)
R2_abso_adult_J <- 1 - SS_res / SS_tot

idx <- complete.cases(abso_test$adult_biomass_TF, pred_abso_pJ, abso_test$season, abso_clim_test)
y_test   <- abso_test$adult_biomass_TF[idx]
y_pred   <- pred_abso_pJ[idx]
clim_use <- abso_clim_test[idx]
SS_res <- sum((y_test - y_pred)^2)
SS_tot <- sum((y_test - clim_use)^2)
R2_abso_adult_pJ <- 1 - SS_res / SS_tot

idx <- complete.cases(anom_test$adult_biomass_TF, pred_anom_J, anom_test$season, anom_clim_test)
y_test   <- anom_test$adult_biomass_TF[idx]
y_pred   <- pred_anom_J[idx]
clim_use <- anom_clim_test[idx]
SS_res <- sum((y_test - y_pred)^2)
SS_tot <- sum((y_test - clim_use)^2)
R2_anom_adult_J <- 1 - SS_res / SS_tot

idx <- complete.cases(anom_test$adult_biomass_TF, pred_anom_pJ, anom_test$season, anom_clim_test)
y_test   <- anom_test$adult_biomass_TF[idx]
y_pred   <- pred_anom_pJ[idx]
clim_use <- anom_clim_test[idx]
SS_res <- sum((y_test - y_pred)^2)
SS_tot <- sum((y_test - clim_use)^2)
R2_anom_adult_pJ <- 1 - SS_res / SS_tot

print(R2_abso_adult_J)
print(R2_abso_adult_pJ)
print(R2_anom_adult_J)
print(R2_anom_adult_pJ)

```

