---
  title: "Do GAM to predict Skipjack Tuna biomass from key environmental variables"
output: html_document
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

For quantifying the effect size and direction, we consider a Generalized Additive Model (GAM) approach. This allows us to interpret the coefficients and explanatory power of the parameters, including linear and nonlinear effects.

```{r setup, include=FALSE}

## 1) remove all objects (incl. hidden)
rm(list = ls(all.names = TRUE))
## 2) unload all non-base packages
pkgs <- names(sessionInfo()$otherPkgs)
if (length(pkgs)) sapply(paste0("package:", pkgs),
                         detach, character.only = TRUE,
                         unload = TRUE, force = TRUE)
## 3) clear plots, console, connections, memory
cat("\014")              # clear console in RStudio
gc()


```

```{r}

setwd("/g/data/es60/SEAPODYM/pearse_GAMs")
list.files()

```

Load the tabulated data

```{r}

my_imfs <- read.table("/g/data/es60/SEAPODYM/pearse_GAMs/forcings_minus_IMFs_seasons.csv", header = TRUE, sep = ",")
my_imfs

# correct the O2
my_imfs$O2_L1 <- my_imfs$O2_L1 / 44.661
my_imfs$O2_L2 <- my_imfs$O2_L2 / 44.661
my_imfs$O2_L3 <- my_imfs$O2_L3 / 44.661
my_imfs$O2_L1_noMF <- my_imfs$O2_L1_noMF / 44.661
my_imfs$O2_L2_noMF <- my_imfs$O2_L2_noMF / 44.661
my_imfs$O2_L3_noMF <- my_imfs$O2_L3_noMF / 44.661
my_imfs$O2_L1_noHF <- my_imfs$O2_L1_noHF / 44.661
my_imfs$O2_L2_noHF <- my_imfs$O2_L2_noHF / 44.661
my_imfs$O2_L3_noHF <- my_imfs$O2_L3_noHF / 44.661
my_imfs$O2_L1_noLF <- my_imfs$O2_L1_noLF / 44.661
my_imfs$O2_L2_noLF <- my_imfs$O2_L2_noLF / 44.661
my_imfs$O2_L3_noLF <- my_imfs$O2_L3_noLF / 44.661
my_imfs$O2_L1_noF <- my_imfs$O2_L1_noF / 44.661
my_imfs$O2_L2_noF <- my_imfs$O2_L2_noF / 44.661
my_imfs$O2_L3_noF <- my_imfs$O2_L3_noF / 44.661


```

Load libraries

```{r}

library(mgcv)
library(parallel)

```

Create cluster for parallel

```{r}

if (exists("cl")) try(stopCluster(cl), silent = TRUE)

cl <- makeCluster(56)      # or a bit fewer to leave headroom
on.exit(stopCluster(cl), add = TRUE)
length(cl)                 # should be 56


```

Fit the GAM model for predicting adults

```{r}

anom <- read.table("/g/data/es60/SEAPODYM/pearse_GAMs/anomalous_values_SEAPODYM_WCPO.txt", header = TRUE, sep = ",")

anom$larvae_biomass_TF <- log10(abs(anom$larvae_biomass)+1.0) * sign(anom$larvae_biomass)
anom$juvenile_biomass_TF <- log10(abs(anom$juvenile_biomass)+1.0) * sign(anom$juvenile_biomass)
anom$adult_biomass_TF <- log10(abs(anom$adult_biomass)+1.0) * sign(anom$adult_biomass)

```

GAM to predict adult skipjack tuna

```{r}


k1 <- 5

### baseline
anom$season_num <- match(anom$season, c("DJF", "MAM", "JJA", "SON"))
anom$season_num <- ifelse(anom$season == "DJF", 1,
                  ifelse(anom$season == "MAM", 2,
                  ifelse(anom$season == "JJA", 3,
                  ifelse(anom$season == "SON", 4, NA))))


```

```{r}

gam_anom_larvaeTF <- bam(
  larvae_biomass_TF ~
    s(pp, k = k1, bs = "ts") +
    s(sst, k = k1, bs = "ts") +
    s(T_L1, k = k1, bs = "ts") +
    s(T_L2, k = k1, bs = "ts") +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 2
)
summary(gam_anom_larvaeTF)

gam_anom_juvenileTF <- bam(
  juvenile_biomass_TF ~
    s(pp, k = k1, bs = "ts") +
    s(sst, k = k1, bs = "ts") +
    s(T_L1, k = k1, bs = "ts") +
    s(T_L2, k = k1, bs = "ts") +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 2
)
summary(gam_anom_juvenileTF)


gam_anom_adultTF <- bam(
  adult_biomass_TF ~
    s(T_L1, k = k1, bs = "ts") +
    s(sst, k = k1, bs = "ts") +
    s(O2_L2, k = k1, bs = "ts") +
    s(T_L2, k = k1, bs = "ts") +
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 2
)
summary(gam_anom_adultTF)


```

Create the anomalies required for the GAM in the my_imfs dataset

```{r}

## vector of base variable names you have IMFs for:
## adjust this to what you actually have
#vars_larvae <- c("pp", "sst", "T_L1", "O2_L2")  # <- add more here
#vars_juvenile <- c("pp", "sst", "T_L1", "O2_L2")  # <- add more here
#vars_adult <- c("T_L1", "sst", "O2_L2", "T_L2")  # <- add more here
vars <- c("pp", "sst", "T_L1", "T_L2", "O2_L2")  # <- add more here

# time index (same for all rows)
t <- seq_len(nrow(my_imfs))

for (v in vars) {
  # column names
  v_noHF <- paste0(v, "_noHF")
  v_noMF <- paste0(v, "_noMF")
  v_noLF <- paste0(v, "_noLF")

  v_HF   <- paste0(v, "_HF")
  v_MF   <- paste0(v, "_MF")
  v_LF   <- paste0(v, "_LF")
  v_T    <- paste0(v, "_T")
  v_Td   <- paste0(v, "_T_detrended")
  v_Tc   <- paste0(v, "_T_centered")

  ## HF/MF/LF from “noX” versions
  my_imfs[[v_HF]] <- my_imfs[[v]] - my_imfs[[v_noHF]]
  my_imfs[[v_MF]] <- my_imfs[[v]] - my_imfs[[v_noMF]]
  my_imfs[[v_LF]] <- my_imfs[[v]] - my_imfs[[v_noLF]]

  ## total residual T
  my_imfs[[v_T]] <- my_imfs[[v]] -
    (my_imfs[[v_MF]] + my_imfs[[v_LF]] + my_imfs[[v_HF]])
  
  my_imfs[[v_Tc]] <- my_imfs[[v_T]] - ave(my_imfs[[v_T]], my_imfs$lon, my_imfs$lat, FUN = function(z) mean(z, na.rm = TRUE))
  
  ## detrend T (linear trend) – NA-safe
  x <- my_imfs[[v_T]]
  ok <- !is.na(x)
  trend <- rep(NA_real_, length(x))
  trend[ok] <- lm(x[ok] ~ t[ok])$fitted.values

  my_imfs[[v_Td]] <- x - trend

}


```


```{r}

lon0 <- 210.5
lat0 <- -2.5

sub <- my_imfs[my_imfs$lon == lon0 & my_imfs$lat == lat0, ]

plot(sub$T_L1_T, type='l')
plot(sub$sst_T, type='l')
plot(sub$O2_L2_T, type='l')
plot(sub$pp_T, type='l')
plot(sub$T_L2_T, type='l')

plot(sub$T_L1_T_detrended, type='l')
plot(sub$sst_T_detrended, type='l')
plot(sub$O2_L2_T_detrended, type='l')
plot(sub$pp_T_detrended, type='l')
plot(sub$T_L2_T_detrended, type='l')

plot(sub$T_L1_T_centered, type='l')
plot(sub$sst_T_centered, type='l')
plot(sub$O2_L2_T_centered, type='l')
plot(sub$pp_T_centered, type='l')
plot(sub$T_L2_T_centered, type='l')


```

Feed the HF, MF, LF and T_centered into GAM


```{r}

cols_to_drop <- c("sst","sst_noMF","sst_noHF","sst_noLF","sst_noVLF","sst_noF",
                  "mld","mld_noMF","mld_noHF","mld_noLF","mld_noVLF","mld_noF",
                  "zeu","zeu_noMF","zeu_noHF","zeu_noLF","zeu_noVLF","zeu_noF",
                  "pp","pp_noMF","pp_noHF","pp_noLF","pp_noVLF","pp_noF",
                  "pp75","pp75_noMF","pp75_noHF","pp75_noLF","pp75_noVLF","pp75_noF",
                  "uo","uo_noMF","uo_noHF","uo_noLF","uo_noVLF","uo_noF",
                  "vo","vo_noMF","vo_noHF","vo_noLF","vo_noVLF","vo_noF",
                  "U_L1","U_L1_noMF","U_L1_noHF","U_L1_noLF","U_L1_noVLF","U_L1_noF",
                  "U_L2","U_L2_noMF","U_L2_noHF","U_L2_noLF","U_L2_noVLF","U_L2_noF",
                  "U_L3","U_L3_noMF","U_L3_noHF","U_L3_noLF","U_L3_noVLF","U_L3_noF",
                  "V_L1","V_L1_noMF","V_L1_noHF","V_L1_noLF","V_L1_noVLF","V_L1_noF",
                  "V_L2","V_L2_noMF","V_L2_noHF","V_L2_noLF","V_L2_noVLF","V_L2_noF",
                  "V_L3","V_L3_noMF","V_L3_noHF","V_L3_noLF","V_L3_noVLF","V_L3_noF",
                  "T_L1","T_L1_noMF","T_L1_noHF","T_L1_noLF","T_L1_noVLF","T_L1_noF",
                  "T_L2","T_L2_noMF","T_L2_noHF","T_L2_noLF","T_L2_noVLF","T_L2_noF",
                  "T_L3","T_L3_noMF","T_L3_noHF","T_L3_noLF","T_L3_noVLF","T_L3_noF",
                  "O2_L1","O2_L1_noMF","O2_L1_noHF","O2_L1_noLF","O2_L1_noVLF","O2_L1_noF",
                  "O2_L2","O2_L2_noMF","O2_L2_noHF","O2_L2_noLF","O2_L2_noVLF","O2_L2_noF",
                  "O2_L3","O2_L3_noMF","O2_L3_noHF","O2_L3_noLF","O2_L3_noVLF","O2_L3_noF")
df_imfs <- my_imfs[, !(names(my_imfs) %in% cols_to_drop), drop = FALSE]

cols_to_drop <- c("T_L1_MF","T_L1_LF","T_L1_T","T_L1_T_detrended","T_L1_T_centered",
                  "sst_MF","sst_LF","sst_T","sst_T_detrended","sst_T_centered",
                  "O2_L2_MF","O2_L2_LF","O2_L2_T","O2_L2_T_detrended","O2_L2_T_centered",
                  "pp_MF","pp_LF","pp_T","pp_T_detrended","pp_T_centered",
                  "T_L2_MF","T_L2_LF","T_L2_T","T_L2_T_detrended","T_L2_T_centered")
df_HF_imfs <- df_imfs[, !(names(df_imfs) %in% cols_to_drop), drop = FALSE]
names(df_HF_imfs) <- sub("_HF$", "", names(df_HF_imfs))

cols_to_drop <- c("T_L1_HF","T_L1_LF","T_L1_T","T_L1_T_detrended","T_L1_T_centered",
                  "sst_HF","sst_LF","sst_T","sst_T_detrended","sst_T_centered",
                  "O2_L2_HF","O2_L2_LF","O2_L2_T","O2_L2_T_detrended","O2_L2_T_centered",
                  "pp_HF","pp_LF","pp_T","pp_T_detrended","pp_T_centered",
                  "T_L2_HF","T_L2_LF","T_L2_T","T_L2_T_detrended","T_L2_T_centered")
df_MF_imfs <- df_imfs[, !(names(df_imfs) %in% cols_to_drop), drop = FALSE]
names(df_MF_imfs) <- sub("_MF$", "", names(df_MF_imfs))

cols_to_drop <- c("T_L1_HF","T_L1_MF","T_L1_T","T_L1_T_detrended","T_L1_T_centered",
                  "sst_HF","sst_MF","sst_T","sst_T_detrended","sst_T_centered",
                  "O2_L2_HF","O2_L2_MF","O2_L2_T","O2_L2_T_detrended","O2_L2_T_centered",
                  "pp_HF","pp_MF","pp_T","pp_T_detrended","pp_T_centered",
                  "T_L2_HF","T_L2_MF","T_L2_T","T_L2_T_detrended","T_L2_T_centered")
df_LF_imfs <- df_imfs[, !(names(df_imfs) %in% cols_to_drop), drop = FALSE]
names(df_LF_imfs) <- sub("_LF$", "", names(df_LF_imfs))

cols_to_drop <- c("T_L1_HF","T_L1_MF","T_L1_T","T_L1_T_detrended","T_L1_LF",
                  "sst_HF","sst_MF","sst_T","sst_T_detrended","sst_LF",
                  "O2_L2_HF","O2_L2_MF","O2_L2_T","O2_L2_T_detrended","O2_L2_LF",
                  "pp_HF","pp_MF","pp_T","pp_T_detrended","pp_LF",
                  "T_L2_HF","T_L2_MF","T_L2_T","T_L2_T_detrended","T_L2_LF")
df_T_imfs <- df_imfs[, !(names(df_imfs) %in% cols_to_drop), drop = FALSE]
names(df_T_imfs) <- sub("_T_centered$", "", names(df_T_imfs))


```

Check that the forcings are correct

```{r}

lon0 <- 210.5
lat0 <- -2.5

sub <- df_T_imfs[df_T_imfs$lon == lon0 & df_T_imfs$lat == lat0, ]

plot(sub$T_L1, type='l')
plot(sub$sst, type='l')
plot(sub$O2_L2, type='l')
plot(sub$pp, type='l')
plot(sub$T_L2, type='l')

```

Do the predictions with the GAM using High-Freq, Mid-Freq, Low-Freq and long-term trend anomalies!

```{r}

### Larvae ###

p <- predict(
  gam_anom_larvaeTF,
  newdata  = df_HF_imfs,
  type     = "response",
  se.fit   = TRUE,
  cluster  = cl,
  block.size = 100000
)
larvaeTF_HF <- p$fit
larvaeTF_HF_se <- p$se.fit


p <- predict(
  gam_anom_larvaeTF,
  newdata  = df_MF_imfs,
  type     = "response",
  se.fit   = TRUE,
  cluster  = cl,
  block.size = 100000
)
larvaeTF_MF <- p$fit
larvaeTF_MF_se <- p$se.fit


p <- predict(
  gam_anom_larvaeTF,
  newdata  = df_LF_imfs,
  type     = "response",
  se.fit   = TRUE,
  cluster  = cl,
  block.size = 100000
)
larvaeTF_LF <- p$fit
larvaeTF_LF_se <- p$se.fit


p <- predict(
  gam_anom_larvaeTF,
  newdata  = df_T_imfs,
  type     = "response",
  se.fit   = TRUE,
  cluster  = cl,
  block.size = 100000
)
larvaeTF_T <- p$fit
larvaeTF_T_se <- p$se.fit


### Juveniles ###

p <- predict(
  gam_anom_juvenileTF,
  newdata  = df_HF_imfs,
  type     = "response",
  se.fit   = TRUE,
  cluster  = cl,
  block.size = 100000
)
juvenileTF_HF <- p$fit
juvenileTF_HF_se <- p$se.fit


p <- predict(
  gam_anom_juvenileTF,
  newdata  = df_MF_imfs,
  type     = "response",
  se.fit   = TRUE,
  cluster  = cl,
  block.size = 100000
)
juvenileTF_MF <- p$fit
juvenileTF_MF_se <- p$se.fit


p <- predict(
  gam_anom_juvenileTF,
  newdata  = df_LF_imfs,
  type     = "response",
  se.fit   = TRUE,
  cluster  = cl,
  block.size = 100000
)
juvenileTF_LF <- p$fit
juvenileTF_LF_se <- p$se.fit


p <- predict(
  gam_anom_juvenileTF,
  newdata  = df_T_imfs,
  type     = "response",
  se.fit   = TRUE,
  cluster  = cl,
  block.size = 100000
)
juvenileTF_T <- p$fit
juvenileTF_T_se <- p$se.fit


### Adults ###

p <- predict(
  gam_anom_adultTF,
  newdata  = df_HF_imfs,
  type     = "response",
  se.fit   = TRUE,
  cluster  = cl,
  block.size = 100000  # try 50k–200k; omit if your mgcv doesn’t have this arg
)
adultTF_HF <- p$fit
adultTF_HF_se  <- p$se.fit


p <- predict(
  gam_anom_adultTF,
  newdata  = df_MF_imfs,
  type     = "response",
  se.fit   = TRUE,
  cluster  = cl,
  block.size = 100000  # try 50k–200k; omit if your mgcv doesn’t have this arg
)
adultTF_MF <- p$fit
adultTF_MF_se  <- p$se.fit


p <- predict(
  gam_anom_adultTF,
  newdata  = df_LF_imfs,
  type     = "response",
  se.fit   = TRUE,
  cluster  = cl,
  block.size = 100000  # try 50k–200k; omit if your mgcv doesn’t have this arg
)
adultTF_LF <- p$fit
adultTF_LF_se  <- p$se.fit


p <- predict(
  gam_anom_adultTF,
  newdata  = df_T_imfs,
  type     = "response",
  se.fit   = TRUE,
  cluster  = cl,
  block.size = 100000  # try 50k–200k; omit if your mgcv doesn’t have this arg
)
adultTF_T <- p$fit
adultTF_T_se  <- p$se.fit

```


```{r}

df_HF_imfs$larvaeTF_HF    <- larvaeTF_HF
df_HF_imfs$larvaeTF_HF_se <- larvaeTF_HF_se
df_HF_imfs$juvenileTF_HF    <- juvenileTF_HF
df_HF_imfs$juvenileTF_HF_se <- juvenileTF_HF_se
df_HF_imfs$adultTF_HF    <- adultTF_HF
df_HF_imfs$adultTF_HF_se <- adultTF_HF_se

df_MF_imfs$larvaeTF_MF    <- larvaeTF_MF
df_MF_imfs$larvaeTF_MF_se <- larvaeTF_MF_se
df_MF_imfs$juvenileTF_MF    <- juvenileTF_MF
df_MF_imfs$juvenileTF_MF_se <- juvenileTF_MF_se
df_MF_imfs$adultTF_MF    <- adultTF_MF
df_MF_imfs$adultTF_MF_se <- adultTF_MF_se

df_LF_imfs$larvaeTF_LF    <- larvaeTF_LF
df_LF_imfs$larvaeTF_LF_se <- larvaeTF_LF_se
df_LF_imfs$juvenileTF_LF    <- juvenileTF_LF
df_LF_imfs$juvenileTF_LF_se <- juvenileTF_LF_se
df_LF_imfs$adultTF_LF    <- adultTF_LF
df_LF_imfs$adultTF_LF_se <- adultTF_LF_se

df_T_imfs$larvaeTF_T    <- larvaeTF_T
df_T_imfs$larvaeTF_T_se <- larvaeTF_T_se
df_T_imfs$juvenileTF_T    <- juvenileTF_T
df_T_imfs$juvenileTF_T_se <- juvenileTF_T_se
df_T_imfs$adultTF_T    <- adultTF_T
df_T_imfs$adultTF_T_se <- adultTF_T_se

nrow(df_HF_imfs); nrow(df_MF_imfs); nrow(df_LF_imfs); nrow(df_T_imfs)

```

```{r}

decompose_predictors <- function(mod, newdata, cluster = NULL, block.size = 100000) {

  # term-level contributions on the linear predictor scale
  pt <- mgcv::predict.gam(
    mod,
    newdata  = newdata,
    type     = "terms",
    se.fit   = FALSE,
    cluster  = cluster,
    block.size = block.size
  )
  term_names <- colnames(pt)

  # helper: sum all terms that contain a given pattern in their name
  sum_by_pattern <- function(pattern) {
    idx <- grep(pattern, term_names)
    if (length(idx) == 0L) return(rep(0, nrow(newdata)))
    rowSums(pt[, idx, drop = FALSE])
  }

  # one column per oceanographic predictor
  contrib_T_L1  <- sum_by_pattern("T_L1")
  contrib_sst   <- sum_by_pattern("sst")
  contrib_O2_L2 <- sum_by_pattern("O2_L2")
  contrib_pp    <- sum_by_pattern("pp")

  # intercept (same for all rows)
  intercept <- attr(pt, "constant")
  if (is.null(intercept) || length(intercept) == 0L) intercept <- 0

  drop <- c(
    "larvaeTF_T_L1",
    "larvaeTF_sst",
    "larvaeTF_O2_L2",
    "larvaeTF_pp",
    "larvaeTF_intercept"
  )
  newdata[drop] <- NULL

  new_cols <- data.frame(
    larvaeTF_T_L1      = contrib_T_L1,
    larvaeTF_sst       = contrib_sst,
    larvaeTF_O2_L2     = contrib_O2_L2,
    larvaeTF_pp        = contrib_pp,
    larvaeTF_intercept = rep(intercept, nrow(newdata)),
    check.names = FALSE
  )

  out <- cbind(newdata, new_cols)
  return(out)
}

# Add predictor-specific contributions
df_HF_imfs <- decompose_predictors(gam_anom_larvaeTF, df_HF_imfs, cluster = cl)
df_MF_imfs <- decompose_predictors(gam_anom_larvaeTF, df_MF_imfs, cluster = cl)
df_LF_imfs <- decompose_predictors(gam_anom_larvaeTF, df_LF_imfs, cluster = cl)
df_T_imfs  <- decompose_predictors(gam_anom_larvaeTF, df_T_imfs, cluster = cl)

```


```{r}

write.csv(df_HF_imfs, "/g/data/es60/SEAPODYM/pearse_GAMs/anomalous_values_GAM_predictions_larvaeHF.csv", row.names = FALSE)
write.csv(df_MF_imfs, "/g/data/es60/SEAPODYM/pearse_GAMs/anomalous_values_GAM_predictions_larvaeMF.csv", row.names = FALSE)
write.csv(df_LF_imfs, "/g/data/es60/SEAPODYM/pearse_GAMs/anomalous_values_GAM_predictions_larvaeLF.csv", row.names = FALSE)
write.csv(df_T_imfs, "/g/data/es60/SEAPODYM/pearse_GAMs/anomalous_values_GAM_predictions_larvaeT.csv", row.names = FALSE)

```


```{r}

decompose_predictors <- function(mod, newdata, cluster = NULL, block.size = 100000) {

  # term-level contributions on the linear predictor scale
  pt <- mgcv::predict.gam(
    mod,
    newdata  = newdata,
    type     = "terms",
    se.fit   = FALSE,
    cluster  = cluster,
    block.size = block.size
  )
  term_names <- colnames(pt)

  # helper: sum all terms that contain a given pattern in their name
  sum_by_pattern <- function(pattern) {
    idx <- grep(pattern, term_names)
    if (length(idx) == 0L) return(rep(0, nrow(newdata)))
    rowSums(pt[, idx, drop = FALSE])
  }

  # one column per oceanographic predictor
  contrib_T_L1  <- sum_by_pattern("T_L1")
  contrib_sst   <- sum_by_pattern("sst")
  contrib_O2_L2 <- sum_by_pattern("O2_L2")
  contrib_pp    <- sum_by_pattern("pp")

  # intercept (same for all rows)
  intercept <- attr(pt, "constant")
  if (is.null(intercept) || length(intercept) == 0L) intercept <- 0

  drop <- c(
    "juvenileTF_T_L1",
    "juvenileTF_sst",
    "juvenileTF_O2_L2",
    "juvenileTF_pp",
    "juvenileTF_intercept"
  )
  newdata[drop] <- NULL

  new_cols <- data.frame(
    juvenileTF_T_L1      = contrib_T_L1,
    juvenileTF_sst       = contrib_sst,
    juvenileTF_O2_L2     = contrib_O2_L2,
    juvenileTF_pp        = contrib_pp,
    juvenileTF_intercept = rep(intercept, nrow(newdata)),
    check.names = FALSE
  )

  out <- cbind(newdata, new_cols)
  return(out)
}

# Add predictor-specific contributions
df_HF_imfs <- decompose_predictors(gam_anom_juvenileTF, df_HF_imfs, cluster = cl)
df_MF_imfs <- decompose_predictors(gam_anom_juvenileTF, df_MF_imfs, cluster = cl)
df_LF_imfs <- decompose_predictors(gam_anom_juvenileTF, df_LF_imfs, cluster = cl)
df_T_imfs  <- decompose_predictors(gam_anom_juvenileTF, df_T_imfs, cluster = cl)

```

```{r}

# --- Save to disk ---
write.csv(df_HF_imfs, "/g/data/es60/SEAPODYM/pearse_GAMs/anomalous_values_GAM_predictions_juvenileHF.csv", row.names = FALSE)
write.csv(df_MF_imfs, "/g/data/es60/SEAPODYM/pearse_GAMs/anomalous_values_GAM_predictions_juvenileMF.csv", row.names = FALSE)
write.csv(df_LF_imfs, "/g/data/es60/SEAPODYM/pearse_GAMs/anomalous_values_GAM_predictions_juvenileLF.csv", row.names = FALSE)
write.csv(df_T_imfs, "/g/data/es60/SEAPODYM/pearse_GAMs/anomalous_values_GAM_predictions_juvenileT.csv", row.names = FALSE)

```

```{r}

decompose_predictors <- function(mod, newdata, cluster = NULL, block.size = 100000) {

  # term-level contributions on the linear predictor scale
  pt <- mgcv::predict.gam(
    mod,
    newdata  = newdata,
    type     = "terms",
    se.fit   = FALSE,
    cluster  = cluster,
    block.size = block.size
  )
  term_names <- colnames(pt)

  # helper: sum all terms that contain a given pattern in their name
  sum_by_pattern <- function(pattern) {
    idx <- grep(pattern, term_names)
    if (length(idx) == 0L) return(rep(0, nrow(newdata)))
    rowSums(pt[, idx, drop = FALSE])
  }

  # one column per oceanographic predictor
  contrib_T_L1 <- sum_by_pattern("T_L1")
  contrib_sst  <- sum_by_pattern("sst")
  contrib_O2_L2<- sum_by_pattern("O2_L2")
  contrib_T_L2 <- sum_by_pattern("T_L2")

  # intercept (same for all rows)
  intercept <- attr(pt, "constant")
  if (is.null(intercept) || length(intercept) == 0L) intercept <- 0

  drop <- c(
    "adultTF_T_L1",
    "adultTF_sst",
    "adultTF_O2_L2",
    "adultTF_T_L2",
    "adultTF_intercept"
    )
  newdata[drop] <- NULL

  new_cols <- data.frame(
    adultTF_T_L1  = contrib_T_L1,
    adultTF_sst   = contrib_sst,
    adultTF_O2_L2 = contrib_O2_L2,
    adultTF_T_L2  = contrib_T_L2,
    adultTF_intercept = rep(intercept, nrow(newdata)),
    check.names = FALSE
  )
  
  out <- cbind(newdata, new_cols)
  return(out)
}

# Add predictor-specific contributions
df_HF_imfs <- decompose_predictors(gam_anom_adultTF, df_HF_imfs, cluster = cl)
df_MF_imfs <- decompose_predictors(gam_anom_adultTF, df_MF_imfs, cluster = cl)
df_LF_imfs <- decompose_predictors(gam_anom_adultTF, df_LF_imfs, cluster = cl)
df_T_imfs  <- decompose_predictors(gam_anom_adultTF, df_T_imfs,  cluster = cl)

```

```{r}

# --- Save to disk ---
write.csv(df_HF_imfs, "/g/data/es60/SEAPODYM/pearse_GAMs/anomalous_values_GAM_predictions_adultHF.csv", row.names = FALSE)
write.csv(df_MF_imfs, "/g/data/es60/SEAPODYM/pearse_GAMs/anomalous_values_GAM_predictions_adultMF.csv", row.names = FALSE)
write.csv(df_LF_imfs, "/g/data/es60/SEAPODYM/pearse_GAMs/anomalous_values_GAM_predictions_adultLF.csv", row.names = FALSE)
write.csv(df_T_imfs, "/g/data/es60/SEAPODYM/pearse_GAMs/anomalous_values_GAM_predictions_adultT.csv", row.names = FALSE)

```

