---
  title: "Do GAM to predict Skipjack Tuna biomass from key environmental variables"
output: html_document
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

For quantifying the effect size and direction, we consider a Generalized Additive Model (GAM) approach. This allows us to interpret the coefficients and explanatory power of the parameters, including linear and nonlinear effects.

```{r setup, include=FALSE}

## 1) remove all objects (incl. hidden)
rm(list = ls(all.names = TRUE))
## 2) unload all non-base packages
pkgs <- names(sessionInfo()$otherPkgs)
if (length(pkgs)) sapply(paste0("package:", pkgs),
                         detach, character.only = TRUE,
                         unload = TRUE, force = TRUE)
## 3) clear plots, console, connections, memory
cat("\014")              # clear console in RStudio
gc()


```

```{r}

setwd("/g/data/es60/SEAPODYM/pearse_GAMs")
list.files()

```

Load the dataframes and transform biomass

```{r}

abso <- read.table("/g/data/es60/SEAPODYM/pearse_GAMs/absolute_values_SEAPODYM_WCPO.txt", header = TRUE, sep = ",")
anom <- read.table("/g/data/es60/SEAPODYM/pearse_GAMs/anomalous_values_SEAPODYM_WCPO.txt", header = TRUE, sep = ",")

abso$larvae_biomass_TF <- log10(abs(abso$larvae_biomass)+1.0) * sign(abso$larvae_biomass)
anom$larvae_biomass_TF <- log10(abs(anom$larvae_biomass)+1.0) * sign(anom$larvae_biomass)

abso$juvenile_biomass_TF <- log10(abs(abso$juvenile_biomass)+1.0) * sign(abso$juvenile_biomass)
anom$juvenile_biomass_TF <- log10(abs(anom$juvenile_biomass)+1.0) * sign(anom$juvenile_biomass)

abso$adult_biomass_TF <- log10(abs(abso$adult_biomass)+1.0) * sign(abso$adult_biomass)
anom$adult_biomass_TF <- log10(abs(anom$adult_biomass)+1.0) * sign(anom$adult_biomass)

```

Load libraries

```{r}

library(mgcv)
library(parallel)

```

Check assumptions of linearity for larvae

```{r}

lm_abso_larvae <- lm(larvae_biomass ~ sst + mld + zeu + pp + uo + vo + U_L1 + U_L2 + U_L3 + V_L1 + V_L2 + V_L3 + O2_L1 + O2_L2 + O2_L3 + T_L1 + T_L2 + T_L3, data=abso)
summary(lm_abso_larvae)

lm_anom_larvae <- lm(larvae_biomass ~ sst + mld + zeu + pp + uo + vo + U_L1 + U_L2 + U_L3 + V_L1 + V_L2 + V_L3 + O2_L1 + O2_L2 + O2_L3 + T_L1 + T_L2 + T_L3, data=anom)
summary(lm_anom_larvae)

png("figures/lm_abso_larvae_diagnostics.png", width = 1600, height = 1200, res = 200, type = "cairo")
op <- par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
plot(lm_abso_larvae, pch = ".", cex = 0.3, id.n = 0)
par(op); dev.off()

png("figures/lm_anom_larvae_diagnostics.png", width = 1600, height = 1200, res = 200, type = "cairo")
op <- par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
plot(lm_anom_larvae, pch = ".", cex = 0.3, id.n = 0)
par(op); dev.off()


```

Raw larval biomass (absolute): R2 = 0.61
Raw larval biomass (anomaly): R2 = 0.47

Try a transformation of the biomass

```{r}

lm_abso_larvaeTF <- lm(larvae_biomass_TF ~ sst + mld + zeu + pp + uo + vo + U_L1 + U_L2 + U_L3 + V_L1 + V_L2 + V_L3 + O2_L1 + O2_L2 + O2_L3 + T_L1 + T_L2 + T_L3, data=abso)
summary(lm_abso_larvaeTF)

lm_anom_larvaeTF <- lm(larvae_biomass_TF ~ sst + mld + zeu + pp + uo + vo + U_L1 + U_L2 + U_L3 + V_L1 + V_L2 + V_L3 + O2_L1 + O2_L2 + O2_L3 + T_L1 + T_L2 + T_L3, data=anom)
summary(lm_anom_larvaeTF)

png("figures/lm_abso_larvaeTF_diagnostics.png", width = 1600, height = 1200, res = 200, type = "cairo")
op <- par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
plot(lm_abso_larvaeTF, pch = ".", cex = 0.3, id.n = 0)
par(op); dev.off()

png("figures/lm_anom_larvaeTF_diagnostics.png", width = 1600, height = 1200, res = 200, type = "cairo")
op <- par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
plot(lm_anom_larvaeTF, pch = ".", cex = 0.3, id.n = 0)
par(op); dev.off()

```

+1 Log10 transformed larval biomass (absolute): R2 = 0.92
+1 Log10 transformed larval biomass (anomaly): R2 = 0.64

Juveniles

```{r}


lm_abso_juvenileTF <- lm(juvenile_biomass_TF ~ sst + mld + zeu + pp + uo + vo + U_L1 + U_L2 + U_L3 + V_L1 + V_L2 + V_L3 + O2_L1 + O2_L2 + O2_L3 + T_L1 + T_L2 + T_L3, data=abso)
summary(lm_abso_juvenileTF)

lm_anom_juvenileTF <- lm(juvenile_biomass_TF ~ sst + mld + zeu + pp + uo + vo + U_L1 + U_L2 + U_L3 + V_L1 + V_L2 + V_L3 + O2_L1 + O2_L2 + O2_L3 + T_L1 + T_L2 + T_L3, data=anom)
summary(lm_anom_juvenileTF)

png("figures/lm_abso_juvenileTF_diagnostics.png", width = 1600, height = 1200, res = 200, type = "cairo")
op <- par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
plot(lm_abso_juvenileTF, pch = ".", cex = 0.3, id.n = 0)
par(op); dev.off()

png("figures/lm_anom_juvenileTF_diagnostics.png", width = 1600, height = 1200, res = 200, type = "cairo")
op <- par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
plot(lm_anom_juvenileTF, pch = ".", cex = 0.3, id.n = 0)
par(op); dev.off()


```

+1.0 Log10 transformed juvenile biomass (absolute): R2 = 0.84
+1.0 Log10 transformed juvenile biomass (anomaly): R2 = 0.42

Adults


```{r}

lm_abso_adultTF <- lm(adult_biomass_TF ~ sst + mld + zeu + pp + uo + vo + U_L1 + U_L2 + U_L3 + V_L1 + V_L2 + V_L3 + O2_L1 + O2_L2 + O2_L3 + T_L1 + T_L2 + T_L3, data=abso)
summary(lm_abso_adultTF)

lm_anom_adultTF <- lm(adult_biomass_TF ~ sst + mld + zeu + pp + uo + vo + U_L1 + U_L2 + U_L3 + V_L1 + V_L2 + V_L3 + O2_L1 + O2_L2 + O2_L3 + T_L1 + T_L2 + T_L3, data=anom)
summary(lm_anom_adultTF)

png("figures/lm_abso_adultTF_diagnostics.png", width = 1600, height = 1200, res = 200, type = "cairo")
op <- par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
plot(lm_abso_adultTF, pch = ".", cex = 0.3, id.n = 0)
par(op); dev.off()

png("figures/lm_anom_adultTF_diagnostics.png", width = 1600, height = 1200, res = 200, type = "cairo")
op <- par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
plot(lm_anom_adultTF, pch = ".", cex = 0.3, id.n = 0)
par(op); dev.off()

```

+1.0 Log10 transformed adult biomass (absolute): R2 = 0.42
+1.0 Log10 transformed adult biomass (anomaly): R2 = 0.08



Create cluster for parallel

```{r}

if (exists("cl")) try(stopCluster(cl), silent = TRUE)

cl <- makeCluster(56)      # or a bit fewer to leave headroom
on.exit(stopCluster(cl), add = TRUE)
length(cl)                 # should be 56


```



```{r}

k1 <- 5

gam_abso_larvaeTF_all <- bam(
  larvae_biomass_TF ~ 
    # main effects (nonlinear, global)
    s(sst, k=k1, bs="ts") + 
    s(mld, k=k1, bs="ts") +
    s(zeu, k=k1, bs="ts") +
    s(uo, k=k1, bs="ts") +
    s(vo, k=k1, bs="ts") +
    s(pp, k=k1, bs="ts") +
    s(T_L1, k=k1, bs="ts") +
    s(T_L2, k=k1, bs="ts") +
    s(T_L3, k=k1, bs="ts") +
    s(U_L1, k=k1, bs="ts") +
    s(U_L2, k=k1, bs="ts") +
    s(U_L3, k=k1, bs="ts") +
    s(V_L1, k=k1, bs="ts") +
    s(V_L2, k=k1, bs="ts") +
    s(V_L3, k=k1, bs="ts") +
    s(O2_L1, k=k1, bs="ts") +
    s(O2_L2, k=k1, bs="ts") +
    s(O2_L3, k=k1, bs="ts") +
    # space-varying effects
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(mld, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(zeu, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(uo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(vo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = abso,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam_abso_larvaeTF_all)

gam_anom_larvaeTF_all <- bam(
  larvae_biomass_TF ~ 
    # main effects (nonlinear, global)
    s(sst, k=k1, bs="ts") + 
    s(mld, k=k1, bs="ts") +
    s(zeu, k=k1, bs="ts") +
    s(uo, k=k1, bs="ts") +
    s(vo, k=k1, bs="ts") +
    s(pp, k=k1, bs="ts") +
    s(T_L1, k=k1, bs="ts") +
    s(T_L2, k=k1, bs="ts") +
    s(T_L3, k=k1, bs="ts") +
    s(U_L1, k=k1, bs="ts") +
    s(U_L2, k=k1, bs="ts") +
    s(U_L3, k=k1, bs="ts") +
    s(V_L1, k=k1, bs="ts") +
    s(V_L2, k=k1, bs="ts") +
    s(V_L3, k=k1, bs="ts") +
    s(O2_L1, k=k1, bs="ts") +
    s(O2_L2, k=k1, bs="ts") +
    s(O2_L3, k=k1, bs="ts") +
    # space-varying effects
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(mld, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(zeu, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(uo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(vo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam_anom_larvaeTF_all)

```

GAM of transformed larval biomass (absolute) with all oceanographic predictors: R2 = 0.978
GAM of transformed larval biomass (anomaly) with all oceanographic predictors: R2 = 0.756

Save GAMs

```{r}

saveRDS(gam_abso_larvaeTF_all, file = "./gams/gam_abso_larvaeTF_all.rds")
saveRDS(gam_anom_larvaeTF_all, file = "./gams/gam_anom_larvaeTF_all.rds")


```

Load GAMs

```{r}

gam_abso_larvaeTF_all <- readRDS("./gams/gam_abso_larvaeTF_all.rds")
gam_anom_larvaeTF_all <- readRDS("./gams/gam_anom_larvaeTF_all.rds")

```

Juveniles

```{r}

gam_abso_juvenileTF_all <- bam(
  juvenile_biomass_TF ~ 
    # main effects (nonlinear, global)
    s(sst, k=k1, bs="ts") + 
    s(mld, k=k1, bs="ts") +
    s(zeu, k=k1, bs="ts") +
    s(uo, k=k1, bs="ts") +
    s(vo, k=k1, bs="ts") +
    s(pp, k=k1, bs="ts") +
    s(T_L1, k=k1, bs="ts") +
    s(T_L2, k=k1, bs="ts") +
    s(T_L3, k=k1, bs="ts") +
    s(U_L1, k=k1, bs="ts") +
    s(U_L2, k=k1, bs="ts") +
    s(U_L3, k=k1, bs="ts") +
    s(V_L1, k=k1, bs="ts") +
    s(V_L2, k=k1, bs="ts") +
    s(V_L3, k=k1, bs="ts") +
    s(O2_L1, k=k1, bs="ts") +
    s(O2_L2, k=k1, bs="ts") +
    s(O2_L3, k=k1, bs="ts") +
    # space-varying effects
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(mld, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(zeu, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(uo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(vo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = abso,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam_abso_juvenileTF_all)

gam_abso_juvenileTF_allL <- bam(
  juvenile_biomass_TF ~ 
    # main effects (nonlinear, global)
    s(sst, k=k1, bs="ts") + 
    s(mld, k=k1, bs="ts") +
    s(zeu, k=k1, bs="ts") +
    s(uo, k=k1, bs="ts") +
    s(vo, k=k1, bs="ts") +
    s(pp, k=k1, bs="ts") +
    s(T_L1, k=k1, bs="ts") +
    s(T_L2, k=k1, bs="ts") +
    s(T_L3, k=k1, bs="ts") +
    s(U_L1, k=k1, bs="ts") +
    s(U_L2, k=k1, bs="ts") +
    s(U_L3, k=k1, bs="ts") +
    s(V_L1, k=k1, bs="ts") +
    s(V_L2, k=k1, bs="ts") +
    s(V_L3, k=k1, bs="ts") +
    s(O2_L1, k=k1, bs="ts") +
    s(O2_L2, k=k1, bs="ts") +
    s(O2_L3, k=k1, bs="ts") +
    s(larvae_biomass_TF, k=k1, bs="ts") +
    # space-varying effects
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(mld, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(zeu, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(uo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(vo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(larvae_biomass_TF, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = abso,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam_abso_juvenileTF_allL)

gam_anom_juvenileTF_all <- bam(
  juvenile_biomass_TF ~ 
    # main effects (nonlinear, global)
    s(sst, k=k1, bs="ts") + 
    s(mld, k=k1, bs="ts") +
    s(zeu, k=k1, bs="ts") +
    s(uo, k=k1, bs="ts") +
    s(vo, k=k1, bs="ts") +
    s(pp, k=k1, bs="ts") +
    s(T_L1, k=k1, bs="ts") +
    s(T_L2, k=k1, bs="ts") +
    s(T_L3, k=k1, bs="ts") +
    s(U_L1, k=k1, bs="ts") +
    s(U_L2, k=k1, bs="ts") +
    s(U_L3, k=k1, bs="ts") +
    s(V_L1, k=k1, bs="ts") +
    s(V_L2, k=k1, bs="ts") +
    s(V_L3, k=k1, bs="ts") +
    s(O2_L1, k=k1, bs="ts") +
    s(O2_L2, k=k1, bs="ts") +
    s(O2_L3, k=k1, bs="ts") +
    # space-varying effects
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(mld, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(zeu, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(uo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(vo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam_anom_juvenileTF_all)

gam_anom_juvenileTF_allL <- bam(
  juvenile_biomass_TF ~ 
    # main effects (nonlinear, global)
    s(sst, k=k1, bs="ts") + 
    s(mld, k=k1, bs="ts") +
    s(zeu, k=k1, bs="ts") +
    s(uo, k=k1, bs="ts") +
    s(vo, k=k1, bs="ts") +
    s(pp, k=k1, bs="ts") +
    s(T_L1, k=k1, bs="ts") +
    s(T_L2, k=k1, bs="ts") +
    s(T_L3, k=k1, bs="ts") +
    s(U_L1, k=k1, bs="ts") +
    s(U_L2, k=k1, bs="ts") +
    s(U_L3, k=k1, bs="ts") +
    s(V_L1, k=k1, bs="ts") +
    s(V_L2, k=k1, bs="ts") +
    s(V_L3, k=k1, bs="ts") +
    s(O2_L1, k=k1, bs="ts") +
    s(O2_L2, k=k1, bs="ts") +
    s(O2_L3, k=k1, bs="ts") +
    s(larvae_biomass_TF, k=k1, bs="ts") +
    # space-varying effects
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(mld, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(zeu, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(uo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(vo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(larvae_biomass_TF, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam_anom_juvenileTF_allL)


```

GAM of transformed juvenile biomass (absolute) with all oceanographic predictors: R2 = 0.94
GAM of transformed juvenile biomass (absolute) with all oceanographic predictors + larvae: R2 = 0.97
GAM of transformed juvenile biomass (anomaly) with all oceanographic predictors: R2 = 0.53
GAM of transformed juvenile biomass (anomaly) with all oceanographic predictors + larvae: R2 = 0.69


```{r}

saveRDS(gam_abso_juvenileTF_all, file = "gam_abso_juvenileTF_all.rds")
saveRDS(gam_abso_juvenileTF_allL, file = "gam_abso_juvenileTF_allL.rds")
saveRDS(gam_anom_juvenileTF_all, file = "gam_anom_juvenileTF_all.rds")
saveRDS(gam_anom_juvenileTF_allL, file = "gam_anom_juvenileTF_allL.rds")

```

Load GAMs

```{r}

gam_abso_juvenileTF_all <- readRDS("gam_abso_juvenileTF_all.rds")
gam_abso_juvenileTF_allL <- readRDS("gam_abso_juvenileTF_allL.rds")
gam_anom_juvenileTF_all <- readRDS("gam_anom_juvenileTF_all.rds")
gam_anom_juvenileTF_allL <- readRDS("gam_anom_juvenileTF_allL.rds")


```



Juveniles using predicted larvae

```{r}

abso$larvae_biomass_TFpred_all <- predict(gam_abso_larvaeTF_all, newdata=abso)
anom$larvae_biomass_TFpred_all <- predict(gam_anom_larvaeTF_all, newdata=anom)

```

```{r}

gam_abso_juvenileTF_allpL <- bam(
  juvenile_biomass_TF ~ 
    # main effects (nonlinear, global)
    s(sst, k=k1, bs="ts") + 
    s(mld, k=k1, bs="ts") +
    s(zeu, k=k1, bs="ts") +
    s(uo, k=k1, bs="ts") +
    s(vo, k=k1, bs="ts") +
    s(pp, k=k1, bs="ts") +
    s(T_L1, k=k1, bs="ts") +
    s(T_L2, k=k1, bs="ts") +
    s(T_L3, k=k1, bs="ts") +
    s(U_L1, k=k1, bs="ts") +
    s(U_L2, k=k1, bs="ts") +
    s(U_L3, k=k1, bs="ts") +
    s(V_L1, k=k1, bs="ts") +
    s(V_L2, k=k1, bs="ts") +
    s(V_L3, k=k1, bs="ts") +
    s(O2_L1, k=k1, bs="ts") +
    s(O2_L2, k=k1, bs="ts") +
    s(O2_L3, k=k1, bs="ts") +
    s(larvae_biomass_TFpred_all, k=k1, bs="ts") +
    # space-varying effects
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(mld, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(zeu, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(uo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(vo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(larvae_biomass_TFpred_all, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = abso,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam_abso_juvenileTF_allpL)


gam_anom_juvenileTF_allpL <- bam(
  juvenile_biomass_TF ~ 
    # main effects (nonlinear, global)
    s(sst, k=k1, bs="ts") + 
    s(mld, k=k1, bs="ts") +
    s(zeu, k=k1, bs="ts") +
    s(uo, k=k1, bs="ts") +
    s(vo, k=k1, bs="ts") +
    s(pp, k=k1, bs="ts") +
    s(T_L1, k=k1, bs="ts") +
    s(T_L2, k=k1, bs="ts") +
    s(T_L3, k=k1, bs="ts") +
    s(U_L1, k=k1, bs="ts") +
    s(U_L2, k=k1, bs="ts") +
    s(U_L3, k=k1, bs="ts") +
    s(V_L1, k=k1, bs="ts") +
    s(V_L2, k=k1, bs="ts") +
    s(V_L3, k=k1, bs="ts") +
    s(O2_L1, k=k1, bs="ts") +
    s(O2_L2, k=k1, bs="ts") +
    s(O2_L3, k=k1, bs="ts") +
    s(larvae_biomass_TFpred_all, k=k1, bs="ts") +
    # space-varying effects
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(mld, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(zeu, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(uo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(vo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(larvae_biomass_TFpred_all, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam_anom_juvenileTF_allpL)


```


GAM of transformed juvenile biomass (absolute) with all oceanographic predictors + predicted larvae: R2 = 0.94
GAM of transformed juvenile biomass (anomaly) with all oceanographic predictors + predicted larvae: R2 = 0.54


```{r}

saveRDS(gam_abso_juvenileTF_allpL, file = "gam_abso_juvenileTF_allpL.rds")
saveRDS(gam_anom_juvenileTF_allpL, file = "gam_anom_juvenileTF_allpL.rds")

```


```{r}

gam_abso_juvenileTF_allpL <- readRDS("gam_abso_juvenileTF_allpL.rds")
gam_anom_juvenileTF_allpL <- readRDS("gam_anom_juvenileTF_allpL.rds")

```


Juveniles ~ larvae

```{r}

gam_abso_juvenileTF_L <- bam(
  juvenile_biomass_TF ~ 
    s(larvae_biomass_TF, k=k1, bs="ts") +
    ti(larvae_biomass_TF, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = abso,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam_abso_juvenileTF_L)

gam_abso_juvenileTF_pL <- bam(
  juvenile_biomass_TF ~ 
    s(larvae_biomass_TFpred_all, k=k1, bs="ts") +
    ti(larvae_biomass_TFpred_all, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = abso,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam_abso_juvenileTF_pL)

gam_anom_juvenileTF_L <- bam(
  juvenile_biomass_TF ~ 
    s(larvae_biomass_TF, k=k1, bs="ts") +
    ti(larvae_biomass_TF, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam_anom_juvenileTF_L)

gam_anom_juvenileTF_pL <- bam(
  juvenile_biomass_TF ~ 
    s(larvae_biomass_TFpred_all, k=k1, bs="ts") +
    ti(larvae_biomass_TFpred_all, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam_anom_juvenileTF_pL)


```


GAM of transformed juvenile biomass (absolute) with larvae: R2 = 0.94
GAM of transformed juvenile biomass (absolute) with predicted larvae from all oceanography: R2 = 0.90
GAM of transformed juvenile biomass (anomaly) with larvae: R2 = 0.64
GAM of transformed juvenile biomass (anomaly) with predicted larvae from all oceanography: R2 = 0.49

Adults

```{r}

gam_abso_adultTF_all <- bam(
  adult_biomass_TF ~ 
    # main effects (nonlinear, global)
    s(sst, k=k1, bs="ts") + 
    s(mld, k=k1, bs="ts") +
    s(zeu, k=k1, bs="ts") +
    s(uo, k=k1, bs="ts") +
    s(vo, k=k1, bs="ts") +
    s(pp, k=k1, bs="ts") +
    s(T_L1, k=k1, bs="ts") +
    s(T_L2, k=k1, bs="ts") +
    s(T_L3, k=k1, bs="ts") +
    s(U_L1, k=k1, bs="ts") +
    s(U_L2, k=k1, bs="ts") +
    s(U_L3, k=k1, bs="ts") +
    s(V_L1, k=k1, bs="ts") +
    s(V_L2, k=k1, bs="ts") +
    s(V_L3, k=k1, bs="ts") +
    s(O2_L1, k=k1, bs="ts") +
    s(O2_L2, k=k1, bs="ts") +
    s(O2_L3, k=k1, bs="ts") +
    # space-varying effects
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(mld, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(zeu, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(uo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(vo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = abso,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam_abso_adultTF_all)

gam_abso_adultTF_allJ <- bam(
  adult_biomass_TF ~ 
    # main effects (nonlinear, global)
    s(sst, k=k1, bs="ts") + 
    s(mld, k=k1, bs="ts") +
    s(zeu, k=k1, bs="ts") +
    s(uo, k=k1, bs="ts") +
    s(vo, k=k1, bs="ts") +
    s(pp, k=k1, bs="ts") +
    s(T_L1, k=k1, bs="ts") +
    s(T_L2, k=k1, bs="ts") +
    s(T_L3, k=k1, bs="ts") +
    s(U_L1, k=k1, bs="ts") +
    s(U_L2, k=k1, bs="ts") +
    s(U_L3, k=k1, bs="ts") +
    s(V_L1, k=k1, bs="ts") +
    s(V_L2, k=k1, bs="ts") +
    s(V_L3, k=k1, bs="ts") +
    s(O2_L1, k=k1, bs="ts") +
    s(O2_L2, k=k1, bs="ts") +
    s(O2_L3, k=k1, bs="ts") +
    s(juvenile_biomass_TF, k=k1, bs="ts") +
    # space-varying effects
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(mld, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(zeu, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(uo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(vo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(juvenile_biomass_TF, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = abso,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam_abso_adultTF_allJ)

gam_anom_adultTF_all <- bam(
  adult_biomass_TF ~ 
    # main effects (nonlinear, global)
    s(sst, k=k1, bs="ts") + 
    s(mld, k=k1, bs="ts") +
    s(zeu, k=k1, bs="ts") +
    s(uo, k=k1, bs="ts") +
    s(vo, k=k1, bs="ts") +
    s(pp, k=k1, bs="ts") +
    s(T_L1, k=k1, bs="ts") +
    s(T_L2, k=k1, bs="ts") +
    s(T_L3, k=k1, bs="ts") +
    s(U_L1, k=k1, bs="ts") +
    s(U_L2, k=k1, bs="ts") +
    s(U_L3, k=k1, bs="ts") +
    s(V_L1, k=k1, bs="ts") +
    s(V_L2, k=k1, bs="ts") +
    s(V_L3, k=k1, bs="ts") +
    s(O2_L1, k=k1, bs="ts") +
    s(O2_L2, k=k1, bs="ts") +
    s(O2_L3, k=k1, bs="ts") +
    # space-varying effects
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(mld, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(zeu, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(uo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(vo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam_anom_adultTF_all)

gam_anom_adultTF_allJ <- bam(
  adult_biomass_TF ~ 
    # main effects (nonlinear, global)
    s(sst, k=k1, bs="ts") + 
    s(mld, k=k1, bs="ts") +
    s(zeu, k=k1, bs="ts") +
    s(uo, k=k1, bs="ts") +
    s(vo, k=k1, bs="ts") +
    s(pp, k=k1, bs="ts") +
    s(T_L1, k=k1, bs="ts") +
    s(T_L2, k=k1, bs="ts") +
    s(T_L3, k=k1, bs="ts") +
    s(U_L1, k=k1, bs="ts") +
    s(U_L2, k=k1, bs="ts") +
    s(U_L3, k=k1, bs="ts") +
    s(V_L1, k=k1, bs="ts") +
    s(V_L2, k=k1, bs="ts") +
    s(V_L3, k=k1, bs="ts") +
    s(O2_L1, k=k1, bs="ts") +
    s(O2_L2, k=k1, bs="ts") +
    s(O2_L3, k=k1, bs="ts") +
    s(juvenile_biomass_TF, k=k1, bs="ts") +
    # space-varying effects
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(mld, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(zeu, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(uo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(vo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(juvenile_biomass_TF, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam_anom_adultTF_allJ)


```

GAM of transformed adult biomass (absolute) with all oceanographic predictors: R2 = 0.81
GAM of transformed adult biomass (absolute) with all oceanographic predictors + juveniles: R2 = 0.84
GAM of transformed adult biomass (anomaly) with all oceanographic predictors: R2 = 0.29
GAM of transformed adult biomass (anomaly) with all oceanographic predictors + juveniles: R2 = 0.37


```{r}

saveRDS(gam_abso_adultTF_all, file = "gam_abso_adultTF_all.rds")
saveRDS(gam_abso_adultTF_allJ, file = "gam_abso_adultTF_allJ.rds")
saveRDS(gam_anom_adultTF_all, file = "gam_anom_adultTF_all.rds")
saveRDS(gam_anom_adultTF_allJ, file = "gam_anom_adultTF_allJ.rds")

```

```{r}

gam_abso_adultTF_all <- readRDS("gam_abso_adultTF_all.rds")
gam_abso_adultTF_allJ <- readRDS("gam_abso_adultTF_allJ.rds")
gam_anom_adultTF_all <- readRDS("gam_anom_adultTF_all.rds")
gam_anom_adultTF_allJ <- readRDS("gam_anom_adultTF_allJ.rds")

```

Adults using predicted juveniles

```{r}

abso$juvenile_biomass_TFpred_all <- predict(gam_abso_juvenileTF_allpL, newdata=abso)
anom$juvenile_biomass_TFpred_all <- predict(gam_anom_juvenileTF_allpL, newdata=anom)

```

```{r}

gam_abso_adultTF_allpJ <- bam(
  adult_biomass_TF ~ 
    # main effects (nonlinear, global)
    s(sst, k=k1, bs="ts") + 
    s(mld, k=k1, bs="ts") +
    s(zeu, k=k1, bs="ts") +
    s(uo, k=k1, bs="ts") +
    s(vo, k=k1, bs="ts") +
    s(pp, k=k1, bs="ts") +
    s(T_L1, k=k1, bs="ts") +
    s(T_L2, k=k1, bs="ts") +
    s(T_L3, k=k1, bs="ts") +
    s(U_L1, k=k1, bs="ts") +
    s(U_L2, k=k1, bs="ts") +
    s(U_L3, k=k1, bs="ts") +
    s(V_L1, k=k1, bs="ts") +
    s(V_L2, k=k1, bs="ts") +
    s(V_L3, k=k1, bs="ts") +
    s(O2_L1, k=k1, bs="ts") +
    s(O2_L2, k=k1, bs="ts") +
    s(O2_L3, k=k1, bs="ts") +
    s(juvenile_biomass_TFpred_all, k=k1, bs="ts") +
    # space-varying effects
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(mld, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(zeu, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(uo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(vo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(juvenile_biomass_TFpred_all, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = abso,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam_abso_adultTF_allpJ)

gam_anom_adultTF_allpJ <- bam(
  adult_biomass_TF ~ 
    # main effects (nonlinear, global)
    s(sst, k=k1, bs="ts") + 
    s(mld, k=k1, bs="ts") +
    s(zeu, k=k1, bs="ts") +
    s(uo, k=k1, bs="ts") +
    s(vo, k=k1, bs="ts") +
    s(pp, k=k1, bs="ts") +
    s(T_L1, k=k1, bs="ts") +
    s(T_L2, k=k1, bs="ts") +
    s(T_L3, k=k1, bs="ts") +
    s(U_L1, k=k1, bs="ts") +
    s(U_L2, k=k1, bs="ts") +
    s(U_L3, k=k1, bs="ts") +
    s(V_L1, k=k1, bs="ts") +
    s(V_L2, k=k1, bs="ts") +
    s(V_L3, k=k1, bs="ts") +
    s(O2_L1, k=k1, bs="ts") +
    s(O2_L2, k=k1, bs="ts") +
    s(O2_L3, k=k1, bs="ts") +
    s(juvenile_biomass_TFpred_all, k=k1, bs="ts") +
    # space-varying effects
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(mld, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(zeu, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(uo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(vo, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(U_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(V_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L3, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(juvenile_biomass_TFpred_all, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam_anom_adultTF_allpJ)

```


GAM of transformed adult biomass (absolute) with all oceanographic predictors + predicted juveniles: R2 = 0.82
GAM of transformed adult biomass (anomaly) with all oceanographic predictors + predicted juveniles: R2 = 0.30


```{r}

saveRDS(gam_abso_adultTF_allpJ, file = "gam_abso_adultTF_allpJ.rds")
saveRDS(gam_anom_adultTF_allpJ, file = "gam_anom_adultTF_allpJ.rds")

```


```{r}

gam_abso_adultTF_allpL <- readRDS("gam_abso_adultTF_allpJ.rds")
gam_abso_adultTF_subsetpL <- readRDS("gam_abso_adultTF_subsetpJ.rds")
gam_anom_adultTF_allpL <- readRDS("gam_anom_adultTF_allpJ.rds")
gam_anom_adultTF_subsetpL <- readRDS("gam_anom_adultTF_subsetpJ.rds")

```


Adults ~ juveniles

```{r}

gam_abso_adultTF_L <- bam(
  adult_biomass_TF ~ 
    s(juvenile_biomass_TF, k=k1, bs="ts") +
    ti(juvenile_biomass_TF, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = abso,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam_abso_adultTF_L)

gam_abso_adultTF_pL <- bam(
  adult_biomass_TF ~ 
    s(juvenile_biomass_TFpred_all, k=k1, bs="ts") +
    ti(juvenile_biomass_TFpred_all, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = abso,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam_abso_adultTF_pL)

gam_anom_adultTF_L <- bam(
  adult_biomass_TF ~ 
    s(juvenile_biomass_TF, k=k1, bs="ts") +
    ti(juvenile_biomass_TF, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam_anom_adultTF_L)

gam_anom_adultTF_pL <- bam(
  adult_biomass_TF ~ 
    s(juvenile_biomass_TFpred_all, k=k1, bs="ts") +
    ti(juvenile_biomass_TFpred_all, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam_anom_adultTF_pL)


```


```{r}

models <- list(
  gam_abso_larvaeTF_all,
  gam_abso_juvenileTF_all,
  gam_abso_juvenileTF_allL,
  gam_abso_juvenileTF_allpL,
  gam_abso_adultTF_all,
  gam_abso_adultTF_allJ,
  gam_abso_adultTF_allpJ,
  
  gam_anom_larvaeTF_all,
  gam_anom_juvenileTF_all,
  gam_anom_juvenileTF_allL,
  gam_anom_juvenileTF_allpL,
  gam_anom_adultTF_all,
  gam_anom_adultTF_allJ,
  gam_anom_adultTF_allpJ
)

names(models) <- c(
  "abso_larvae_all", 
  "abso_juv_all", "abso_juv_allL", "abso_juv_allpL",
  "abso_adult_all", "abso_adult_allJ", "abso_adult_allpJ",
  "anom_larvae_all",
  "anom_juv_all", "anom_juv_allL", "anom_juv_allpL",
  "anom_adult_all", "anom_adult_allJ", "anom_adult_allpJ"
)

sapply(models, function(m) format(AIC(m), scientific = TRUE))


```




