---
  title: "Do GAM to predict Skipjack Tuna biomass from key environmental variables"
output: html_document
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

For quantifying the effect size and direction, we consider a Generalized Additive Model (GAM) approach. This allows us to interpret the coefficients and explanatory power of the parameters, including linear and nonlinear effects.

```{r setup, include=FALSE}

## 1) remove all objects (incl. hidden)
rm(list = ls(all.names = TRUE))
## 2) unload all non-base packages
pkgs <- names(sessionInfo()$otherPkgs)
if (length(pkgs)) sapply(paste0("package:", pkgs),
                         detach, character.only = TRUE,
                         unload = TRUE, force = TRUE)
## 3) clear plots, console, connections, memory
cat("\014")              # clear console in RStudio
gc()


```

```{r}

setwd("/g/data/es60/SEAPODYM/pearse_GAMs")
list.files()

```

Load the dataframes and transform biomass

```{r}

abso <- read.table("/g/data/es60/SEAPODYM/pearse_GAMs/absolute_values_SEAPODYM_WCPO.txt", header = TRUE, sep = ",")
anom <- read.table("/g/data/es60/SEAPODYM/pearse_GAMs/anomalous_values_SEAPODYM_WCPO.txt", header = TRUE, sep = ",")

abso$larvae_biomass_TF <- log10(abs(abso$larvae_biomass)+1.0) * sign(abso$larvae_biomass)
anom$larvae_biomass_TF <- log10(abs(anom$larvae_biomass)+1.0) * sign(anom$larvae_biomass)

abso$juvenile_biomass_TF <- log10(abs(abso$juvenile_biomass)+1.0) * sign(abso$juvenile_biomass)
anom$juvenile_biomass_TF <- log10(abs(anom$juvenile_biomass)+1.0) * sign(anom$juvenile_biomass)

abso$adult_biomass_TF <- log10(abs(abso$adult_biomass)+1.0) * sign(abso$adult_biomass)
anom$adult_biomass_TF <- log10(abs(anom$adult_biomass)+1.0) * sign(anom$adult_biomass)

```

Load libraries

```{r}

library(mgcv)
library(parallel)

```


Create cluster for parallel

```{r}

if (exists("cl")) try(stopCluster(cl), silent = TRUE)

cl <- makeCluster(28)      # or a bit fewer to leave headroom
on.exit(stopCluster(cl), add = TRUE)
length(cl)                 # should be 56


```

Load GAMs

```{r}

# Larvae
gam_abso_larvaeTF_all <- readRDS("/g/data/es60/SEAPODYM/pearse_GAMs/gams/gam_abso_larvaeTF.rds")
gam_anom_larvaeTF_all <- readRDS("/g/data/es60/SEAPODYM/pearse_GAMs/gams/gam_anom_larvaeTF.rds")

# Juveniles
gam_abso_juvenileTF_all <- readRDS("/g/data/es60/SEAPODYM/pearse_GAMs/gams/gam_abso_juvenileTF.rds")
gam_abso_juvenileTF_allL <- readRDS("/g/data/es60/SEAPODYM/pearse_GAMs/gams/gam_abso_juvenileTF_L.rds")
gam_anom_juvenileTF_all <- readRDS("/g/data/es60/SEAPODYM/pearse_GAMs/gams/gam_anom_juvenileTF.rds")
gam_anom_juvenileTF_allL <- readRDS("/g/data/es60/SEAPODYM/pearse_GAMs/gams/gam_anom_juvenileTF_L.rds")
gam_abso_juvenileTF_allpL <- readRDS("/g/data/es60/SEAPODYM/pearse_GAMs/gams/gam_abso_juvenileTF_pL.rds")
gam_anom_juvenileTF_allpL <- readRDS("/g/data/es60/SEAPODYM/pearse_GAMs/gams/gam_anom_juvenileTF_pL.rds")

# Adults
gam_abso_adultTF_all <- readRDS("/g/data/es60/SEAPODYM/pearse_GAMs/gams/gam_abso_adultTF.rds")
gam_abso_adultTF_allJ <- readRDS("/g/data/es60/SEAPODYM/pearse_GAMs/gams/gam_abso_adultTF_J.rds")
gam_anom_adultTF_all <- readRDS("/g/data/es60/SEAPODYM/pearse_GAMs/gams/gam_anom_adultTF.rds")
gam_anom_adultTF_allJ <- readRDS("/g/data/es60/SEAPODYM/pearse_GAMs/gams/gam_anom_adultTF_J.rds")
gam_abso_adultTF_allpL <- readRDS("/g/data/es60/SEAPODYM/pearse_GAMs/gams/gam_abso_adultTF_pJ.rds")
gam_anom_adultTF_allpL <- readRDS("/g/data/es60/SEAPODYM/pearse_GAMs/gams/gam_anom_adultTF_pJ.rds")

```



Build up a model from scratch, trying each of the individual predictors within the subset, and record R2

1. start with a baseline that tries to predict anomalous tuna biomass from cyclic seasonality and space

```{r}

k1 <- 5

### baseline
anom$season_num <- match(anom$season, c("DJF", "MAM", "JJA", "SON"))
anom$season_num <- ifelse(anom$season == "DJF", 1,
                  ifelse(anom$season == "MAM", 2,
                  ifelse(anom$season == "JJA", 3,
                  ifelse(anom$season == "SON", 4, NA))))


gam_anom_larvaeTF_baseline <- bam(
  larvae_biomass_TF ~
    # tamed spatial baseline
    s(lon,lat, bs="ds", k = 200) +
    # seasonality
    s(season_num, bs="cc", k=4),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam_anom_larvaeTF_baseline)

gam_anom_juvenileTF_baseline <- bam(
  juvenile_biomass_TF ~
    # tamed spatial baseline
    s(lon,lat, bs="ds", k = 200) +
    # seasonality
    s(season_num, bs="cc", k=4),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam_anom_juvenileTF_baseline)

gam_anom_adultTF_baseline <- bam(
  adult_biomass_TF ~
    # tamed spatial baseline
    s(lon,lat, bs="ds", k = 200) +
    # seasonality
    s(season_num, bs="cc", k=4),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 1
)
summary(gam_anom_adultTF_baseline)

```


Larval absolute (lat,lon, season) = 71.3%
Juvenile absolute (lat,lon, season) = 71.2%
Adult absolute (lat,lon, season) = 64.8%


Larval predictor importance, one at a time

SST, NPP

```{r}


k1 <- 5

gam_anom_larvaeTF_sst <- update(gam_anom_larvaeTF_baseline,
                       . ~ . + s(sst, k = k1, bs = "ts") + ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2))
summary(gam_anom_larvaeTF_sst)

gam_anom_larvaeTF_pp <- update(gam_anom_larvaeTF_baseline,
                       . ~ . + s(pp, k = k1, bs = "ts") + ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2))
summary(gam_anom_larvaeTF_pp)

gam_anom_larvaeTF_T_L1 <- update(gam_anom_larvaeTF_baseline,
                       . ~ . + s(T_L1, k = k1, bs = "ts") + ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2))
summary(gam_anom_larvaeTF_T_L1)

gam_anom_larvaeTF_T_L2 <- update(gam_anom_larvaeTF_baseline,
                       . ~ . + s(T_L2, k = k1, bs = "ts") + ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2))
summary(gam_anom_larvaeTF_T_L2)

gam_anom_larvaeTF_U_L1 <- update(gam_anom_larvaeTF_baseline,
                       . ~ . + s(U_L1, k = k1, bs = "ts") + ti(U_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2))
summary(gam_anom_larvaeTF_U_L1)

gam_anom_larvaeTF_U_L2 <- update(gam_anom_larvaeTF_baseline,
                       . ~ . + s(U_L2, k = k1, bs = "ts") + ti(U_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2))
summary(gam_anom_larvaeTF_U_L2)

gam_anom_larvaeTF_V_L1 <- update(gam_anom_larvaeTF_baseline,
                       . ~ . + s(V_L1, k = k1, bs = "ts") + ti(V_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2))
summary(gam_anom_larvaeTF_V_L1)

gam_anom_larvaeTF_V_L2 <- update(gam_anom_larvaeTF_baseline,
                       . ~ . + s(V_L2, k = k1, bs = "ts") + ti(V_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2))
summary(gam_anom_larvaeTF_V_L2)

gam_anom_larvaeTF_O2_L1 <- update(gam_anom_larvaeTF_baseline,
                       . ~ . + s(O2_L1, k = k1, bs = "ts") + ti(O2_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2))
summary(gam_anom_larvaeTF_O2_L1)

gam_anom_larvaeTF_O2_L2 <- update(gam_anom_larvaeTF_baseline,
                       . ~ . + s(O2_L2, k = k1, bs = "ts") + ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2))
summary(gam_anom_larvaeTF_O2_L2)

```


Make functions for local seasonal climatology and skill assessment against this

```{r}

make_climatology <- function(data, response) {
  tmp <- data.frame(lon = data$lon, lat = data$lat, season = data$season, y = data[[response]])
  clim <- aggregate(y ~ lon + lat + season, data = tmp, FUN = mean, na.rm = TRUE)
  names(clim)[4] <- "clim"
  clim
}

clim_skill <- function(model, eval_data, climatology, response) {
  pred <- predict(model, newdata = eval_data, type = "response")
  key_clim <- paste(climatology$lon, climatology$lat, climatology$season, sep = "_")
  key_eval <- paste(eval_data$lon, eval_data$lat, eval_data$season, sep = "_")
  clim_pred <- climatology$clim[ match(key_eval, key_clim) ]
  obs <- eval_data[[response]]
  idx <- complete.cases(obs, pred, clim_pred)
  SS_res <- sum((obs[idx] - pred[idx])^2)
  SS_clim <- sum((obs[idx] - clim_pred[idx])^2)
  1 - SS_res / SS_clim
}

```

Do forward selection for larvae:

```{r}

k1 <- 5

# vars: character vector of predictor names
# k1:   your chosen k for the 1D smooths
build_terms <- function(vars, k1) {
  if (length(vars) == 0L) return(character(0))

  g_terms  <- sprintf("s(%s, k=%s, bs='ts')", vars, k1)
  ti_terms <- sprintf(
    "ti(%s, lon, lat, bs=c('cr','tp','tp'), k=c(%s,120), m=2)",
    vars, k1
  )

  c(g_terms, ti_terms)  # return all terms as a character vector
}

forward_gam_stepwise <- function(predictors,
                                 preselected = NULL,   # e.g. "T_L1"
                                 train_data,
                                 eval_data,
                                 baseline_terms,
                                 response,
                                 k1,
                                 thresh = 0.01,        # 1 percentage point
                                 verbose = TRUE,
                                 ...) {

  ## ---- set up selected / remaining ----
  selected  <- unique(preselected)
  remaining <- setdiff(predictors, selected)

  ## Climatology calculated ONLY from training data
  climatology <- make_climatology(train_data, response)

  ## ---- fit baseline (+ preselected, if any) ----
  rhs_terms <- c(baseline_terms, build_terms(selected, k1))
  rhs <- paste(rhs_terms, collapse = " + ")
  base_formula <- as.formula(paste(response, "~", rhs))

  if (verbose) {
    cat("Fitting starting model with:\n  ",
        if (length(selected)) paste(selected, collapse = ", ") else "baseline only",
        "\n")
  }

  best_model <- bam(base_formula, data = train_data, ...)
  current_dev <- summary(best_model)$dev.expl
  current_skill <- clim_skill(best_model, eval_data, climatology, response)

  if (verbose)
    cat("Starting climatology skill:", round(current_skill, 3), "\n")

  trace <- data.frame(
    step     = integer(0),
    added    = character(0),
    skill    = numeric(0),
    delta    = numeric(0),
    stringsAsFactors = FALSE
  )

  step <- 0L

  ## ---- forward selection loop ----
  repeat {
    if (length(remaining) == 0L) {
      if (verbose) cat("No predictors left.\n")
      break
    }

    step <- step + 1L
    if (verbose)
      cat("\nStep", step, "- testing", length(remaining),
          "candidate predictors\n")

    best_var_step <- NULL
    best_skill_step <- current_skill
    best_model_step <- NULL

    ## test each remaining predictor
    for (v in remaining) {
      rhs_terms_v <- c(
        baseline_terms,
        build_terms(selected, k1),
        build_terms(v, k1)
      )
      rhs_v <- paste(rhs_terms_v, collapse = " + ")
      f_v   <- as.formula(paste(response, "~", rhs_v))

      m_v <- bam(f_v, data = train_data, ...)

      dev_v <- summary(m_v)$dev.expl
      skill_v <- clim_skill(m_v, eval_data, climatology, response)

      if (verbose)
        cat("  ", v, "-> skill =", round(skill_v, 3), "\n")

      if (!is.na(skill_v) && skill_v > best_skill_step) {
        best_skill_step <- skill_v
        best_var_step <- v
        best_model_step <- m_v
      }

      ## free memory from this candidate
      rm(m_v); gc()
    }

    ## nothing improved?
    delta <- best_skill_step - current_skill

    if (is.null(best_var_step) || delta < thresh) {

      if (verbose) {
        cat("Best improvement <", thresh, "(Δ skill =", round(delta, 3),"). Stopping.\n")
      }
      break
    }

    ## accept the best candidate
    selected    <- c(selected, best_var_step)
    remaining   <- setdiff(remaining, best_var_step)
    current_skill <- best_skill_step
    best_model  <- best_model_step

    trace <- rbind(trace, data.frame(
      step     = step,
      added    = best_var_step,
      skill    = current_skill,
      delta    = delta,
      stringsAsFactors = FALSE
    ))

    if (verbose) {
      cat("Selected:", best_var_step, "| skill =", round(current_skill, 3), "| Δ =", round(delta, 3), "\n")
    }
    rm(best_model_step); gc()
  }

  list(
    selected    = selected,
    trace       = trace,
    final_model = best_model,
    climatology = climatology,
    validation_skill = current_skill
  )
}

```


```{r}


### baseline
anom$season_num <- match(anom$season, c("DJF", "MAM", "JJA", "SON"))
anom$season_num <- ifelse(anom$season == "DJF", 1,
                  ifelse(anom$season == "MAM", 2,
                  ifelse(anom$season == "JJA", 3,
                  ifelse(anom$season == "SON", 4, NA))))


baseline_terms <- "s(lon,lat,bs='ds',k=200) + s(season_num,bs='cc',k=4)"

predictors <- c(
  "sst","pp",
  "T_L1","T_L2",
  "U_L1","U_L2",
  "V_L1","V_L2",
  "O2_L1","O2_L2"
)

# ============================================================
# 80/20 temporal train-test split
# ============================================================
years <- sort(unique(anom$season_year))
cut_year <- floor(0.8 * length(years))
train_years <- years[1:cut_year]
test_years  <- years[(cut_year + 1):length(years)]
anom_train <- anom[ anom$season_year %in% train_years, ]
anom_test <- anom[ anom$season_year %in% test_years, ]

cat("Training years:", min(train_years), "-", max(train_years), "\n")
cat("Testing years:", min(test_years), "-", max(test_years), "\n")
cat("Training rows:", nrow(anom_train), "\n", "Testing rows:", nrow(anom_test), "\n")

```


```{r}

# ============================================================
# Forward selection
#
# Candidate GAMs are fitted to abso_train.
# Candidate performance is evaluated on abso_test relative
# to the lon × lat × season climatology from abso_train.
# ============================================================

res <- forward_gam_stepwise(
  predictors     = predictors,
  preselected    = "pp",
  train_data     = anom_train,
  eval_data      = anom_test,
  baseline_terms = baseline_terms,
  response       = "larvae_biomass_TF",
  k1             = k1,
  thresh         = 0.01,        # require ≥0.01 increase in climatology skill
  method         = "fREML",
  select         = TRUE,
  cluster        = cl,
  gc.level       = 2
)

res$selected
res$trace
res$validation_skill
summary(res$final_model)


```

Juvenile predictor importance, one at a time

SST, NPP

```{r}


k1 <- 5


gam_anom_juvenileTF_sst <- update(gam_anom_juvenileTF_baseline,
                       . ~ . + s(sst, k = k1, bs = "ts") + ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2))
summary(gam_anom_juvenileTF_sst)

gam_anom_juvenileTF_pp <- update(gam_anom_juvenileTF_baseline,
                       . ~ . + s(pp, k = k1, bs = "ts") + ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2))
summary(gam_anom_juvenileTF_pp)

gam_anom_juvenileTF_T_L1 <- update(gam_anom_juvenileTF_baseline,
                       . ~ . + s(T_L1, k = k1, bs = "ts") + ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2))
summary(gam_anom_juvenileTF_T_L1)

gam_anom_juvenileTF_T_L2 <- update(gam_anom_juvenileTF_baseline,
                       . ~ . + s(T_L2, k = k1, bs = "ts") + ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2))
summary(gam_anom_juvenileTF_T_L2)

gam_anom_juvenileTF_U_L1 <- update(gam_anom_juvenileTF_baseline,
                       . ~ . + s(U_L1, k = k1, bs = "ts") + ti(U_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2))
summary(gam_anom_juvenileTF_U_L1)

gam_anom_juvenileTF_U_L2 <- update(gam_anom_juvenileTF_baseline,
                       . ~ . + s(U_L2, k = k1, bs = "ts") + ti(U_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2))
summary(gam_anom_juvenileTF_U_L2)

gam_anom_juvenileTF_V_L1 <- update(gam_anom_juvenileTF_baseline,
                       . ~ . + s(V_L1, k = k1, bs = "ts") + ti(V_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2))
summary(gam_anom_juvenileTF_V_L1)

gam_anom_juvenileTF_V_L2 <- update(gam_anom_juvenileTF_baseline,
                       . ~ . + s(V_L2, k = k1, bs = "ts") + ti(V_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2))
summary(gam_anom_juvenileTF_V_L2)

gam_anom_juvenileTF_O2_L1 <- update(gam_anom_juvenileTF_baseline,
                       . ~ . + s(O2_L1, k = k1, bs = "ts") + ti(O2_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2))
summary(gam_anom_juvenileTF_O2_L1)

gam_anom_juvenileTF_O2_L2 <- update(gam_anom_juvenileTF_baseline,
                       . ~ . + s(O2_L2, k = k1, bs = "ts") + ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2))
summary(gam_anom_juvenileTF_O2_L2)

```


Juveniles forward selection

```{r}


# ============================================================
# Forward selection
#
# Candidate GAMs are fitted to abso_train.
# Candidate performance is evaluated on abso_test relative
# to the lon × lat × season climatology from abso_train.
# ============================================================

res <- forward_gam_stepwise(
  predictors     = predictors,
  preselected    = "pp",
  train_data     = anom_train,
  eval_data      = anom_test,
  baseline_terms = baseline_terms,
  response       = "juvenile_biomass_TF",
  k1             = k1,
  thresh         = 0.01,        # require ≥0.01 increase in climatology skill
  method         = "fREML",
  select         = TRUE,
  cluster        = cl,
  gc.level       = 2
)

res$selected
res$trace
res$validation_skill
summary(res$final_model)


```

Adult predictor importance, one at a time


```{r}


k1 <- 5

gam_anom_adultTF_sst <- update(gam_anom_adultTF_baseline,
                       . ~ . + s(sst, k = k1, bs = "ts") + ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2))
summary(gam_anom_adultTF_sst)

gam_anom_adultTF_pp <- update(gam_anom_adultTF_baseline,
                       . ~ . + s(pp, k = k1, bs = "ts") + ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2))
summary(gam_anom_adultTF_pp)

gam_anom_adultTF_T_L1 <- update(gam_anom_adultTF_baseline,
                       . ~ . + s(T_L1, k = k1, bs = "ts") + ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2))
summary(gam_anom_adultTF_T_L1)

gam_anom_adultTF_T_L2 <- update(gam_anom_adultTF_baseline,
                       . ~ . + s(T_L2, k = k1, bs = "ts") + ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2))
summary(gam_anom_adultTF_T_L2)

gam_anom_adultTF_U_L1 <- update(gam_anom_adultTF_baseline,
                       . ~ . + s(U_L1, k = k1, bs = "ts") + ti(U_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2))
summary(gam_anom_adultTF_U_L1)

gam_anom_adultTF_U_L2 <- update(gam_anom_adultTF_baseline,
                       . ~ . + s(U_L2, k = k1, bs = "ts") + ti(U_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2))
summary(gam_anom_adultTF_U_L2)

gam_anom_adultTF_V_L1 <- update(gam_anom_adultTF_baseline,
                       . ~ . + s(V_L1, k = k1, bs = "ts") + ti(V_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2))
summary(gam_anom_adultTF_V_L1)

gam_anom_adultTF_V_L2 <- update(gam_anom_adultTF_baseline,
                       . ~ . + s(V_L2, k = k1, bs = "ts") + ti(V_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2))
summary(gam_anom_adultTF_V_L2)

gam_anom_adultTF_O2_L1 <- update(gam_anom_adultTF_baseline,
                       . ~ . + s(O2_L1, k = k1, bs = "ts") + ti(O2_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2))
summary(gam_anom_adultTF_O2_L1)

gam_anom_adultTF_O2_L2 <- update(gam_anom_adultTF_baseline,
                       . ~ . + s(O2_L2, k = k1, bs = "ts") + ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2))
summary(gam_anom_adultTF_O2_L2)


```


```{r}

# ============================================================
# Forward selection
#
# Candidate GAMs are fitted to abso_train.
# Candidate performance is evaluated on abso_test relative
# to the lon × lat × season climatology from abso_train.
# ============================================================

res <- forward_gam_stepwise(
  predictors     = predictors,
  preselected    = "T_L1",
  train_data     = anom_train,
  eval_data      = anom_test,
  baseline_terms = baseline_terms,
  response       = "adult_biomass_TF",
  k1             = k1,
  thresh         = 0.01,        # require ≥0.01 increase in climatology skill
  method         = "fREML",
  select         = TRUE,
  cluster        = cl,
  gc.level       = 2
)

res$selected
res$trace
res$validation_skill
summary(res$final_model)

```


Assess explanatory power of global versus local relationships


```{r}

gam_anom_adultTF_sstG <- update(gam_anom_adultTF_baseline,
                       . ~ . + s(sst, k = k1, bs = "ts"))
summary(gam_anom_adultTF_sstG)

gam_anom_adultTF_sstV <- update(gam_anom_adultTF_baseline,
                       . ~ . + ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2))
summary(gam_anom_adultTF_sstV)

gam_anom_adultTF_sst <- update(gam_anom_adultTF_baseline,
                       . ~ . + s(sst, k = k1, bs = "ts") + ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2))
summary(gam_anom_adultTF_sst)

gam_anom_adultTF_ppG <- update(gam_anom_adultTF_baseline,
                       . ~ . + s(pp, k = k1, bs = "ts"))
summary(gam_anom_adultTF_ppG)

gam_anom_adultTF_ppV <- update(gam_anom_adultTF_baseline,
                       . ~ . + ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2))
summary(gam_anom_adultTF_ppV)

gam_anom_adultTF_pp <- update(gam_anom_adultTF_baseline,
                       . ~ . + s(pp, k = k1, bs = "ts") + ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2))
summary(gam_anom_adultTF_pp)


```


```{r}

gam_anom_larvaeTF_FS1 <- bam(
  larvae_biomass_TF ~
    s(pp, k = k1, bs = "ts") +
    s(sst, k = k1, bs = "ts") +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 2
)
summary(gam_anom_larvaeTF_FS1)

gam_anom_larvaeTF_FS2 <- bam(
  larvae_biomass_TF ~
    s(pp, k = k1, bs = "ts") +
    s(sst, k = k1, bs = "ts") +
    s(T_L1, k = k1, bs = "ts") +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 2
)
summary(gam_anom_larvaeTF_FS2)

gam_anom_larvaeTF_FS3 <- bam(
  larvae_biomass_TF ~
    s(pp, k = k1, bs = "ts") +
    s(sst, k = k1, bs = "ts") +
    s(T_L1, k = k1, bs = "ts") +
    s(O2_L2, k = k1, bs = "ts") +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 2
)
summary(gam_anom_larvaeTF_FS3)

gam_anom_larvaeTF_FS1G <- bam(
  larvae_biomass_TF ~
    s(pp, k = k1, bs = "ts") +
    s(sst, k = k1, bs = "ts"),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 2
)
summary(gam_anom_larvaeTF_FS1G)

gam_anom_larvaeTF_FS2G <- bam(
  larvae_biomass_TF ~
    s(pp, k = k1, bs = "ts") +
    s(sst, k = k1, bs = "ts") +
    s(T_L1, k = k1, bs = "ts"),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 2
)
summary(gam_anom_larvaeTF_FS2G)

gam_anom_larvaeTF_FS3G <- bam(
  larvae_biomass_TF ~
    s(pp, k = k1, bs = "ts") +
    s(sst, k = k1, bs = "ts") +
    s(T_L1, k = k1, bs = "ts") +
    s(O2_L2, k = k1, bs = "ts"),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 2
)
summary(gam_anom_larvaeTF_FS3G)

gam_anom_larvaeTF_FS1V <- bam(
  larvae_biomass_TF ~
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 2
)
summary(gam_anom_larvaeTF_FS1V)

gam_anom_larvaeTF_FS2V <- bam(
  larvae_biomass_TF ~
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 2
)
summary(gam_anom_larvaeTF_FS2V)

gam_anom_larvaeTF_FS3V <- bam(
  larvae_biomass_TF ~
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 2
)
summary(gam_anom_larvaeTF_FS3V)

```


```{r}


gam_anom_juvenileTF_FS1 <- bam(
  juvenile_biomass_TF ~
    s(pp, k = k1, bs = "ts") +
    s(sst, k = k1, bs = "ts") +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 2
)
summary(gam_anom_juvenileTF_FS1)

gam_anom_juvenileTF_FS2 <- bam(
  juvenile_biomass_TF ~
    s(pp, k = k1, bs = "ts") +
    s(sst, k = k1, bs = "ts") +
    s(T_L1, k = k1, bs = "ts") +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 2
)
summary(gam_anom_juvenileTF_FS2)

gam_anom_juvenileTF_FS3 <- bam(
  juvenile_biomass_TF ~
    s(pp, k = k1, bs = "ts") +
    s(sst, k = k1, bs = "ts") +
    s(T_L1, k = k1, bs = "ts") +
    s(O2_L2, k = k1, bs = "ts") +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 2
)
summary(gam_anom_juvenileTF_FS3)


gam_anom_juvenileTF_FS1G <- bam(
  juvenile_biomass_TF ~
    s(pp, k = k1, bs = "ts") +
    s(sst, k = k1, bs = "ts"),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 2
)
summary(gam_anom_juvenileTF_FS1G)

gam_anom_juvenileTF_FS2G <- bam(
  juvenile_biomass_TF ~
    s(pp, k = k1, bs = "ts") +
    s(sst, k = k1, bs = "ts") +
    s(T_L1, k = k1, bs = "ts"),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 2
)
summary(gam_anom_juvenileTF_FS2G)

gam_anom_juvenileTF_FS3G <- bam(
  juvenile_biomass_TF ~
    s(pp, k = k1, bs = "ts") +
    s(sst, k = k1, bs = "ts") +
    s(T_L1, k = k1, bs = "ts") +
    s(O2_L2, k = k1, bs = "ts"),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 2
)
summary(gam_anom_juvenileTF_FS3G)

gam_anom_juvenileTF_FS1V <- bam(
  juvenile_biomass_TF ~
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 2
)
summary(gam_anom_juvenileTF_FS1V)

gam_anom_juvenileTF_FS2V <- bam(
  juvenile_biomass_TF ~
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 2
)
summary(gam_anom_juvenileTF_FS2V)

gam_anom_juvenileTF_FS3V <- bam(
  juvenile_biomass_TF ~
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(pp, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 2
)
summary(gam_anom_juvenileTF_FS3V)


```

```{r}


gam_anom_adultTF_FS1 <- bam(
  adult_biomass_TF ~
    s(sst, k = k1, bs = "ts") +
    s(T_L1, k = k1, bs = "ts") +
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 2
)
summary(gam_anom_adultTF_FS1)

gam_anom_adultTF_FS2 <- bam(
  adult_biomass_TF ~
    s(sst, k = k1, bs = "ts") +
    s(T_L1, k = k1, bs = "ts") +
    s(O2_L2, k = k1, bs = "ts") +
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 2
)
summary(gam_anom_adultTF_FS2)

gam_anom_adultTF_FS3 <- bam(
  adult_biomass_TF ~
    s(sst, k = k1, bs = "ts") +
    s(T_L1, k = k1, bs = "ts") +
    s(T_L2, k = k1, bs = "ts") +
    s(O2_L2, k = k1, bs = "ts") +
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 2
)
summary(gam_anom_adultTF_FS3)

gam_anom_adultTF_FS1G <- bam(
  adult_biomass_TF ~
    s(sst, k = k1, bs = "ts") +
    s(T_L1, k = k1, bs = "ts"),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 2
)
summary(gam_anom_adultTF_FS1G)

gam_anom_adultTF_FS2G <- bam(
  adult_biomass_TF ~
    s(sst, k = k1, bs = "ts") +
    s(T_L1, k = k1, bs = "ts") +
    s(O2_L2, k = k1, bs = "ts"),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 2
)
summary(gam_anom_adultTF_FS2G)

gam_anom_adultTF_FS3G <- bam(
  adult_biomass_TF ~
    s(sst, k = k1, bs = "ts") +
    s(T_L1, k = k1, bs = "ts") +
    s(T_L2, k = k1, bs = "ts") +
    s(O2_L2, k = k1, bs = "ts"),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 2
)
summary(gam_anom_adultTF_FS3G)


gam_anom_adultTF_FS1V <- bam(
  adult_biomass_TF ~
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 2
)
summary(gam_anom_adultTF_FS1V)

gam_anom_adultTF_FS2V <- bam(
  adult_biomass_TF ~
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 2
)
summary(gam_anom_adultTF_FS2V)

gam_anom_adultTF_FS3V <- bam(
  adult_biomass_TF ~
    ti(sst, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L1, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(T_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2) +
    ti(O2_L2, lon, lat, bs=c("cr","tp","tp"), k=c(k1, 120), m=2),
  data = anom,
  method = "fREML",
  select = TRUE,
  cluster = cl,
  gc.level = 2
)
summary(gam_anom_adultTF_FS3V)

```

Save global effects to dataframes

```{r}

save_global_effect <- function(mod, data, var, n = 500, file) {
  # build sequence over observed range
  x_seq <- seq(min(data[[var]], na.rm = TRUE),
               max(data[[var]], na.rm = TRUE),
               length.out = n)

  # template: repeat mean values for n rows
  newdat <- data.frame(
    lon        = rep(mean(data$lon, na.rm = TRUE), n),
    lat        = rep(mean(data$lat, na.rm = TRUE), n),
    season_num = rep(round(mean(data$season_num, na.rm = TRUE)), n),

    pp    = rep(0.0, n),
    sst   = rep(0.0, n),
    T_L1  = rep(0.0, n),
    O2_L2 = rep(0.0, n)
  )

  newdat[[var]] <- x_seq  # vary only this var

  # get all term contributions + SEs
  tt <- predict(mod, newdata = newdat, type = "terms", se.fit = TRUE)

  # find the column name for s(var)
  term_name <- paste0("s(", var, ")")
  if (!term_name %in% colnames(tt$fit)) {
    stop("Term ", term_name, " not found in model terms.")
  }

  out <- data.frame(
    x      = x_seq,
    effect = tt$fit[, term_name],
    se     = tt$se.fit[, term_name]
  )

  write.csv(out, file, row.names = FALSE)
  invisible(out)
}


save_global_effect(gam_anom_larvaeTF_FS3, anom, "pp",  file = "gam_larvae_FS3_global_pp.csv")
save_global_effect(gam_anom_larvaeTF_FS3, anom, "sst",   file = "gam_larvae_FS3_global_sst.csv")
save_global_effect(gam_anom_larvaeTF_FS3, anom, "T_L1", file = "gam_larvae_FS3_global_T_L1.csv")
save_global_effect(gam_anom_larvaeTF_FS3, anom, "O2_L2",  file = "gam_larvae_FS3_global_O2_L2.csv")


```

Save global effects to dataframes

```{r}

save_global_effect <- function(mod, data, var, n = 500, file) {
  # build sequence over observed range
  x_seq <- seq(min(data[[var]], na.rm = TRUE),
               max(data[[var]], na.rm = TRUE),
               length.out = n)

  # template: repeat mean values for n rows
  newdat <- data.frame(
    lon        = rep(mean(data$lon, na.rm = TRUE), n),
    lat        = rep(mean(data$lat, na.rm = TRUE), n),
    season_num = rep(round(mean(data$season_num, na.rm = TRUE)), n),

    pp    = rep(0.0, n),
    sst   = rep(0.0, n),
    T_L1  = rep(0.0, n),
    O2_L2 = rep(0.0, n)
  )

  newdat[[var]] <- x_seq  # vary only this var

  # get all term contributions + SEs
  tt <- predict(mod, newdata = newdat, type = "terms", se.fit = TRUE)

  # find the column name for s(var)
  term_name <- paste0("s(", var, ")")
  if (!term_name %in% colnames(tt$fit)) {
    stop("Term ", term_name, " not found in model terms.")
  }

  out <- data.frame(
    x      = x_seq,
    effect = tt$fit[, term_name],
    se     = tt$se.fit[, term_name]
  )

  write.csv(out, file, row.names = FALSE)
  invisible(out)
}


save_global_effect(gam_anom_juvenileTF_FS3, anom, "pp",  file = "gam_juvenile_FS3_global_pp.csv")
save_global_effect(gam_anom_juvenileTF_FS3, anom, "sst",   file = "gam_juvenile_FS3_global_sst.csv")
save_global_effect(gam_anom_juvenileTF_FS3, anom, "T_L1", file = "gam_juvenile_FS3_global_T_L1.csv")
save_global_effect(gam_anom_juvenileTF_FS3, anom, "O2_L2",  file = "gam_juvenile_FS3_global_O2_L2.csv")


```

```{r}

save_global_effect <- function(mod, data, var, n = 500, file) {
  # build sequence over observed range
  x_seq <- seq(min(data[[var]], na.rm = TRUE),
               max(data[[var]], na.rm = TRUE),
               length.out = n)

  # template: repeat mean values for n rows
  newdat <- data.frame(
    lon        = rep(mean(data$lon, na.rm = TRUE), n),
    lat        = rep(mean(data$lat, na.rm = TRUE), n),
    season_num = rep(round(mean(data$season_num, na.rm = TRUE)), n),

    T_L1  = rep(0.0, n),
    sst   = rep(0.0, n),
    O2_L2 = rep(0.0, n),
    T_L2  = rep(0.0, n)
  )

  newdat[[var]] <- x_seq  # vary only this var

  # get all term contributions + SEs
  tt <- predict(mod, newdata = newdat, type = "terms", se.fit = TRUE)

  # find the column name for s(var)
  term_name <- paste0("s(", var, ")")
  if (!term_name %in% colnames(tt$fit)) {
    stop("Term ", term_name, " not found in model terms.")
  }

  out <- data.frame(
    x      = x_seq,
    effect = tt$fit[, term_name],
    se     = tt$se.fit[, term_name]
  )

  write.csv(out, file, row.names = FALSE)
  invisible(out)
}


save_global_effect(gam_anom_adultTF_FS3, anom, "T_L1",  file = "gam_adult_FS3_global_T_L1.csv")
save_global_effect(gam_anom_adultTF_FS3, anom, "sst",   file = "gam_adult_FS3_global_sst.csv")
save_global_effect(gam_anom_adultTF_FS3, anom, "O2_L2", file = "gam_adult_FS3_global_O2_L2.csv")
save_global_effect(gam_anom_adultTF_FS3, anom, "T_L2",  file = "gam_adult_FS3_global_T_L2.csv")


```


Save the spatial effects

```{r}

save_spatial_effect_sd <- function(mod, data, var,
                                   lon_n   = 110,
                                   lat_n   = 40,
                                   seasons = 1:4,
                                   delta   = NULL,
                                   file) {

  # use actual model grid, not seq(min,max,…)
  lon_seq <- sort(unique(data$lon))
  lat_seq <- sort(unique(data$lat))

  grid <- expand.grid(
    lon        = lon_seq,
    lat        = lat_seq,
    season_num = seasons
  )

  base <- transform(
    grid,
    pp    = 0.0,
    sst   = 0.0,
    T_L1  = 0.0,
    O2_L2 = 0.0
  )

  if (is.null(delta)) {
    # local SD over time per (lon,lat)
    sd_loc <- stats::aggregate(
      data[[var]],
      by = list(lon = data$lon, lat = data$lat),
      FUN = function(z) {
        z <- z[is.finite(z)]
        if (length(z) < 2L) {
          NA_real_                  # land / too few data -> NA
        } else {
          stats::sd(z)
        }
      }
    )
    names(sd_loc)[3] <- "sd_var"

    base <- merge(base, sd_loc, by = c("lon", "lat"), all.x = TRUE)
  } else {
    if (!is.finite(delta) || delta == 0) {
      stop("Non-finite or zero user-specified delta for ", var)
    }
    base$sd_var <- rep(delta, nrow(base))
  }

  base_pred <- predict(mod, newdata = base, type = "response")

  mults <- c(-2, -1, 1, 2)
  out_list <- vector("list", length(mults))

  for (i in seq_along(mults)) {
    m <- mults[i]
    delta_val <- m * base$sd_var       # NA on land, scalar elsewhere

    pert <- base
    pert[[var]] <- pert[[var]] + delta_val

    pred_pert <- predict(mod, newdata = pert, type = "response")

    out_list[[i]] <- data.frame(
      lon         = base$lon,
      lat         = base$lat,
      season_num  = base$season_num,
      delta_mult  = m,
      delta_value = delta_val,
      delta_eff   = pred_pert - base_pred
    )
  }

  out <- do.call(rbind, out_list)
  write.csv(out, file, row.names = FALSE)
  invisible(out)
}

```


```{r}

save_spatial_effect_sd(gam_anom_larvaeTF_FS3, anom, "pp",  file = "gam_larvae_FS3_spatial_pp_sd.csv")
save_spatial_effect_sd(gam_anom_larvaeTF_FS3, anom, "sst",   file = "gam_larvae_FS3_spatial_sst_sd.csv")
save_spatial_effect_sd(gam_anom_larvaeTF_FS3, anom, "T_L1", file = "gam_larvae_FS3_spatial_T_L1_sd.csv")
save_spatial_effect_sd(gam_anom_larvaeTF_FS3, anom, "O2_L2",  file = "gam_larvae_FS3_spatial_O2_L2_sd.csv")

save_spatial_effect_sd(gam_anom_larvaeTF_FS3, anom, "pp",  delta=5, file = "gam_larvae_FS3_spatial_pp_5mmolCm2day.csv")
save_spatial_effect_sd(gam_anom_larvaeTF_FS3, anom, "sst",   delta=0.5, file = "gam_larvae_FS3_spatial_sst_0.5degC.csv")
save_spatial_effect_sd(gam_anom_larvaeTF_FS3, anom, "T_L1", delta=0.5, file = "gam_larvae_FS3_spatial_T_L1_0.5deg.csv")
save_spatial_effect_sd(gam_anom_larvaeTF_FS3, anom, "O2_L2",  delta=10.0/44.661, file = "gam_larvae_FS3_spatial_O2_L2_10mmolm3.csv")

```

```{r}

save_spatial_effect_sd(gam_anom_juvenileTF_FS3, anom, "pp",  file = "gam_juvenile_FS3_spatial_pp_sd.csv")
save_spatial_effect_sd(gam_anom_juvenileTF_FS3, anom, "sst",   file = "gam_juvenile_FS3_spatial_sst_sd.csv")
save_spatial_effect_sd(gam_anom_juvenileTF_FS3, anom, "T_L1", file = "gam_juvenile_FS3_spatial_T_L1_sd.csv")
save_spatial_effect_sd(gam_anom_juvenileTF_FS3, anom, "O2_L2",  file = "gam_juvenile_FS3_spatial_O2_L2_sd.csv")

save_spatial_effect_sd(gam_anom_juvenileTF_FS3, anom, "pp",  delta=5, file = "gam_juvenile_FS3_spatial_pp_5mmolCm2day.csv")
save_spatial_effect_sd(gam_anom_juvenileTF_FS3, anom, "sst",   delta=0.5, file = "gam_juvenile_FS3_spatial_sst_0.5degC.csv")
save_spatial_effect_sd(gam_anom_juvenileTF_FS3, anom, "T_L1", delta=0.5, file = "gam_juvenile_FS3_spatial_T_L1_0.5deg.csv")
save_spatial_effect_sd(gam_anom_juvenileTF_FS3, anom, "O2_L2",  delta=10.0/44.661, file = "gam_juvenile_FS3_spatial_O2_L2_10mmolm3.csv")

```

Save the spatial effects

```{r}

save_spatial_effect_sd <- function(mod, data, var,
                                   lon_n   = 110,
                                   lat_n   = 40,
                                   seasons = 1:4,
                                   delta   = NULL,
                                   file) {

  # use actual model grid, not seq(min,max,…)
  lon_seq <- sort(unique(data$lon))
  lat_seq <- sort(unique(data$lat))

  grid <- expand.grid(
    lon        = lon_seq,
    lat        = lat_seq,
    season_num = seasons
  )

  base <- transform(
    grid,
    T_L1  = 0.0,
    sst   = 0.0,
    O2_L2 = 0.0,
    T_L2  = 0.0
  )

  if (is.null(delta)) {
    # local SD over time per (lon,lat)
    sd_loc <- stats::aggregate(
      data[[var]],
      by = list(lon = data$lon, lat = data$lat),
      FUN = function(z) {
        z <- z[is.finite(z)]
        if (length(z) < 2L) {
          NA_real_                  # land / too few data -> NA
        } else {
          stats::sd(z)
        }
      }
    )
    names(sd_loc)[3] <- "sd_var"

    base <- merge(base, sd_loc, by = c("lon", "lat"), all.x = TRUE)
  } else {
    if (!is.finite(delta) || delta == 0) {
      stop("Non-finite or zero user-specified delta for ", var)
    }
    base$sd_var <- rep(delta, nrow(base))
  }

  base_pred <- predict(mod, newdata = base, type = "response")

  mults <- c(-2, -1, 1, 2)
  out_list <- vector("list", length(mults))

  for (i in seq_along(mults)) {
    m <- mults[i]
    delta_val <- m * base$sd_var       # NA on land, scalar elsewhere

    pert <- base
    pert[[var]] <- pert[[var]] + delta_val

    pred_pert <- predict(mod, newdata = pert, type = "response")

    out_list[[i]] <- data.frame(
      lon         = base$lon,
      lat         = base$lat,
      season_num  = base$season_num,
      delta_mult  = m,
      delta_value = delta_val,
      delta_eff   = pred_pert - base_pred
    )
  }

  out <- do.call(rbind, out_list)
  write.csv(out, file, row.names = FALSE)
  invisible(out)
}

```

```{r}

save_spatial_effect_sd(gam_anom_adultTF_FS3, anom, "T_L1",  file = "gam_adult_FS3_spatial_T_L1_sd.csv")
save_spatial_effect_sd(gam_anom_adultTF_FS3, anom, "sst",   file = "gam_adult_FS3_spatial_sst_sd.csv")
save_spatial_effect_sd(gam_anom_adultTF_FS3, anom, "O2_L2", file = "gam_adult_FS3_spatial_O2_L2_sd.csv")
save_spatial_effect_sd(gam_anom_adultTF_FS3, anom, "T_L2",  file = "gam_adult_FS3_spatial_T_L2_sd.csv")


save_spatial_effect_sd(gam_anom_adultTF_FS3, anom, "T_L1",  delta=0.5, file = "gam_adult_FS3_spatial_T_L1_0.5degC.csv")
save_spatial_effect_sd(gam_anom_adultTF_FS3, anom, "sst",   delta=0.5, file = "gam_adult_FS3_spatial_sst_0.5degC.csv")
save_spatial_effect_sd(gam_anom_adultTF_FS3, anom, "O2_L2", delta=10.0/44.661, file = "gam_adult_FS3_spatial_O2_L2_10mmolm3.csv")
save_spatial_effect_sd(gam_anom_adultTF_FS3, anom, "T_L2",  delta=0.5, file = "gam_adult_FS3_spatial_T_L2_0.5degC.csv")

```


Save the predicted larvae biomass to csv

```{r}

## 1a) Overall prediction
anom$larvae_biomassTF_pred_FS3 <- predict(
  gam_anom_larvaeTF_FS3,
  newdata = anom,
  type    = "response"
)

## 1b) Term-wise contributions (matrix)
terms_mat <- predict(
  gam_anom_larvaeTF_FS3,
  newdata = anom,
  type    = "terms"
)

## Check names once, so you know what you’re adding:
colnames(terms_mat)

anom_export <- cbind(anom, as.data.frame(terms_mat))

anom_export$eff_pp  <- anom_export[,"s(pp)"]  + anom_export[,"ti(pp,lon,lat)"]
anom_export$eff_sst <- anom_export[,"s(sst)"] + anom_export[,"ti(sst,lon,lat)"]
anom_export$eff_T_L1 <- anom_export[,"s(T_L1)"] + anom_export[,"ti(T_L1,lon,lat)"]
anom_export$eff_O2_L2 <- anom_export[,"s(O2_L2)"]  + anom_export[,"ti(O2_L2,lon,lat)"]

write.csv(
  anom_export,
  file      = "anomalous_larvae_GAM_FS3terms_WCPO.csv",
  row.names = FALSE
)

```

Save the predicted juvenile biomass to csv

```{r}

## 1a) Overall prediction
anom$juvenile_biomassTF_pred_FS3 <- predict(
  gam_anom_juvenileTF_FS3,
  newdata = anom,
  type    = "response"
)

## 1b) Term-wise contributions (matrix)
terms_mat <- predict(
  gam_anom_juvenileTF_FS3,
  newdata = anom,
  type    = "terms"
)

## Check names once, so you know what you’re adding:
colnames(terms_mat)

anom_export <- cbind(anom, as.data.frame(terms_mat))

anom_export$eff_pp  <- anom_export[,"s(pp)"]  + anom_export[,"ti(pp,lon,lat)"]
anom_export$eff_sst <- anom_export[,"s(sst)"] + anom_export[,"ti(sst,lon,lat)"]
anom_export$eff_T_L1 <- anom_export[,"s(T_L1)"] + anom_export[,"ti(T_L1,lon,lat)"]
anom_export$eff_O2_L2 <- anom_export[,"s(O2_L2)"]  + anom_export[,"ti(O2_L2,lon,lat)"]

write.csv(
  anom_export,
  file      = "anomalous_juvenile_GAM_FS3terms_WCPO.csv",
  row.names = FALSE
)

```

Save the predicted adult biomass to csv

```{r}

## 1a) Overall prediction
anom$adult_biomassTF_pred_FS3 <- predict(
  gam_anom_adultTF_FS3,
  newdata = anom,
  type    = "response"
)

## 1b) Term-wise contributions (matrix)
terms_mat <- predict(
  gam_anom_adultTF_FS3,
  newdata = anom,
  type    = "terms"
)

## Check names once, so you know what you’re adding:
colnames(terms_mat)

anom_export <- cbind(anom, as.data.frame(terms_mat))

anom_export$eff_sst  <- anom_export[,"s(sst)"]  + anom_export[,"ti(sst,lon,lat)"]
anom_export$eff_T_L1 <- anom_export[,"s(T_L1)"] + anom_export[,"ti(T_L1,lon,lat)"]
anom_export$eff_O2_L2 <- anom_export[,"s(O2_L2)"] + anom_export[,"ti(O2_L2,lon,lat)"]
anom_export$eff_T_L2 <- anom_export[,"s(T_L2)"]  + anom_export[,"ti(T_L2,lon,lat)"]

write.csv(
  anom_export,
  file      = "anomalous_adult_GAM_FS3terms_WCPO.csv",
  row.names = FALSE
)

```

